Pathway activation model for personalized prediction of drug synergy

Version 1

Abstract

Abstract

Targeted monotherapies for cancer often fail due to inherent or acquired drug resistance. By aiming at multiple targets simultaneously, drug combinations can produce synergistic interactions that increase drug effectiveness and reduce resistance. Computational models based on the integration of omics data have been used to identify synergistic combinations, but predicting drug synergy remains a challenge. Here, we introduce DIPx, an algorithm for personalized prediction of drug synergy based on biologically motivated tumor- and drug-specific pathway activation scores (PASs). We trained and validated DIPx in the AstraZeneca-Sanger (AZS) DREAM Challenge dataset using two separate test sets: Test Set 1 comprised the combinations already present in the training set, while Test Set 2 contained combinations absent from the training set, thus indicating the model’s ability to handle novel combinations. The Spearman correlation coefficients between predicted and observed drug synergy were 0.50 (95% CI: 0.47–0.53) in Test Set 1 and 0.26 (95% CI: 0.22–0.30) in Test Set 2, compared to 0.38 (95% CI: 0.34–0.42) and 0.18 (95% CI: 0.16–0.20), respectively, for the best performing method in the Challenge. We show evidence that higher synergy is associated with higher functional interaction between the drug targets, and this functional interaction information is captured by PAS. We illustrate the use of PAS to provide a potential biological explanation in terms of activated pathways that mediate the synergistic effects of combined drugs. In summary, DIPx can be a useful tool for personalized prediction of drug synergy and exploration of activated pathways related to the effects of combined drugs.

Introduction

Targeted therapies such as specific inhibitors are the most promising class of cancer drugs, but often fail or achieve only temporary remission due to inherent or acquired resistance. Theoretically, by aiming at multiple targets simultaneously, drug combinations can produce a synergistic interaction that increases drug effectiveness and reduces resistance and the chances of relapse (Medicine, 2017; Pemovska et al., 2018; Plana et al., 2022). This is illustrated in the combination of a BRAF inhibitor dabrafenib with a MEK inhibitor trametinib, which suppresses paradoxical reactivation and resistance observed in patients with BRAF-mutated melanoma treated with dabrafenib alone (Zhong et al., 2022; Banzi et al., 2016). This recently approved combination has been shown to improve progression-free and overall survival rates (Subbiah et al., 2023).

The discovery of effective drug combinations has traditionally relied on expert knowledge and understanding of known biological mechanisms (Li et al., 2015). However, this expert-based approach has limited scope to come up with novel combinations. Furthermore, ideally, novel combinations are clinically tested, but it is practically impossible to test all reasonable combinations in a clinical setting. Computational models of drug synergy have shown some potential for personalized prediction of synergistic combinations (Güvenç Paltun et al., 2021; Wu et al., 2022; Kong et al., 2022). These models are typically based on the integration of patient-specific molecular data, such as mutation profiles, gene expression, and drug response information (Güvenç Paltun et al., 2021). For example, TAJI, the best performing method in the AstraZeneca-Sanger DREAM Challenge, uses these diverse data types to predict drug synergy (Li et al., 2018). The drug combinations predicted to be effective will expand the therapeutic options while maintaining the same level of adverse effects profile. However, despite the advantages offered by modern machine learning methodologies and the availability of large-scale datasets, prediction of synergistic combinations and validating computational models remains challenging. For example, drug screening protocols often vary across studies, and there is a limited overlap in tested drugs and cell lines, complicating the external validation of these models. Additionally, the reliance on ‘black-box’ machine learning approaches hinders the exploration of underlying molecular mechanisms driving synergistic combinations.

To address this limitation, several studies have introduced statistical and computational approaches to infer the mechanisms of action of synergistic combinations within cancer signaling pathways. For example, Liu et al. proposed TranSynergy, a drug synergy prediction model that uses the interaction between drug target genes in a protein-protein interaction (PPI) network (Liu and Xie, 2021). However, TranSynergy only relies on target gene information, neglecting information on upstream and downstream activities of the targets and their differential contributions to synergy. More recently, Tang et al. developed SynPathy, a deep learning model for drug synergy prediction that incorporates drug-pathway associations (Tang and Gottlieb, 2022). SynPathy calculates pathway enrichment scores as a measure of the distance between target genes of each drug in a combination and pathway genes in the PPI network. These pathway enrichment scores, along with chemical structure information, are then combined to fit the model and infer pathway importance scores for each combination. More recently, Wu et al. introduced ForSyn (Wu et al., 2023), a deep forest-based method. Although ForSyn implemented a gene enrichment analysis to identify cancer-related pathways, it does not directly identify them through prediction.

Here we present a Drug synergy Interaction Prediction (DIPx) based on tumor- and drug-specific pathway activation scores (PASs). PASs are biologically motivated features that provide potentially relevant information on the underlying mechanisms of synergistic combinations. We trained and validated DIPx using the AstraZeneca-Sanger (AZS) DREAM Challenge dataset (Menden et al., 2019), and compared its performance with the best performing method in the Challenge. Furthermore, we assessed the generalizability of the model by validating it on the ONeil dataset (O’Neil et al., 2016), and provided illustrations of pathways that could mediate the synergistic combinations found by DIPx. DIPx is publicly available at https://www.github.com/tracquangthinh/DIPx.

Results

A pathway based drug synergy prediction model

Figure 1 provides an overview of DIPx, which uses gene expression, mutation profiles, and drug synergy data from the AZS dataset to train and validate its prediction model. The test set comprised two subsets: (i) Test Set 1 includes combinations from the training set, and (ii) Test Set 2 includes combinations absent from the training set. Together, both sets assess the generalizability of the prediction for new patients and new combinations. The analysis involved a total of 75 cell lines tested in 910 combinations in the AZS dataset. DIPx was also validated using an external dataset, as shown in Figure 1a.

Overview of DIPx. a) The AZS omics data were used to train and validate the model. The test set was split into two subsets: Test Set 1 contained combinations found in the training set, while Set 2 comprised combinations not found in the training set. The model was also externally validated using the ONeil dataset. b) A cartoon illustration of the ERBB pathway in a breast cancer cell line treated with the combination of Capivasertib + Sapitinib. Capivasertib targets the AKT gene, whereas Sapitinib targets the ERBB genes. Pathway genes were classified into upstream and downstream genes relative to the position of the target genes in the network. c) The drug synergy prediction model was trained using pathway activation scores (PAS) of the upstream, downstream and driver genes. d) The predicted and observed Loewe scores of a cell line achieved the median Spearman correlation in Test Set 1 of the AZS dataset. The color of each bar shows the confidence score information with the threshold of 0.75. e) The main pathways that contribute to the prediction of the synergy of the Capivasertib + Sapitinib combination. f) The proportion of validated high synergistic predictions (Loewe score ≥ 20) increases with higher confidence scores. The x-axis presents four groups defined by quartiles of confidence scores.

Figure 1b illustrates the ERBB signaling pathway in relation to the Capivasertib + Sapitinib combination, where the genes belonging to the pathway are classified into upstream and downstream genes relative to the position of the target genes: AKT targeted by Capivasertib and ERBB targeted by Sapitinib. Putative driver mutations were identified in each sample based on a well-characterized list of frequently mutated genes in cancer; see Section 4.3. We first calculate the PAS of the upstream and downstream part of the pathway relative to the driver genes; see the Methods section for details. The PAS values are then combined to train a random forest regression model. Given a new drug combination experiment, DIPx predicts the Loewe score for drug synergy, as shown in Figure 1c.

Figure 1d presents the predicted and observed synergies for the SW900 lung cancer cell line, which has a median correlation of 0.50 among the cell lines in Test Set 1; each bar in the figure represents a drug combination. The best predicted combinations include BCL2L1 + AZD5582, AZD5582 + etoposide and doxorubicin + AZD5582, with predicted Loewe scores of 42.34, 26.60, and 25.72, respectively, and high confidence scores of 1.0, 0.90, and 0.82, respectively. A combination with Loewe score greater than 20 is considered highly synergistic (Menden et al., 2019). Although the combination of doxorubicin + AZ12623380 is predicted to have high synergy, it is a low confidence prediction with a confidence score of 0.33. Indeed, the observed Loewe synergy score for this combination is near zero.

The use of PAS allows DIPx to infer the potential biological mechanisms of synergistic drug combinations. Figure 1e shows pathways with the highest contribution to prediction of drug synergy of the Capivasertib + Sapitinib combination: these include the ERBB-related pathways (ERBB2 signaling pathway, ERBB pathway), and tumor-related pathway (lymph-node metastates, focal adhesion).

Figure 1f demonstrates the association between the confidence scores and the validation of predictions. The x-axis represents the confidence scores grouped into quartiles, while the y-axis displays the proportion of validated high synergy (Loewe score ≥ 20). Predictions with higher confidence scores are expected to exhibit a greater level of validation. Indeed, in this figure, the proportion of high synergistic predictions that are validated in the combination of Test Set 1 and 2 of the AZS dataset increases as the confidence score rises.

Validation and comparisons in the AZS dataset

We evaluated the performance of DIPx in the AZS test sets and compared it with TAJI, which was the best performing method in the AZS DREAM Challenge (Li et al., 2018). TAJI was trained using both monotherapy drug-response and molecular data. Since DIPx uses only molecular data, to make a fair comparison, we trained TAJI using only molecular features and referred to it as TAJI-M. The extra information from the use of monotherapy data in TAJI is rather small, approximately 10% increase in the overall Spearman correlation, and, of course, we could also use such data in DIPx, so it is more convenient and informative to focus the comparisons on prediction based on molecular data alone. For instance, this allows us to compare DIPx with TAJI-M on the prediction of combinations that contain un-trained drug(s), which is not possible with TAJI.

Figure 2a shows the correlation between the predicted and observed Loewe scores of 963 experiments in Test Set 1 (r = 0.5, 95% CI: 0.47–0.53), where each experiment represents a combination drug A + drug B tried on cell line C, yielding one data point. In comparison, TAJI-M gives r = 0.38 (95% CI: 0.34–0.42). We also bootstrapped the training set (n = 100 times) and for each bootstrap replicate calculated the Spearman correlation between the predicted and observed scores of all experiments. As illustrated in Figure 2b, DIPx achieved stable Spearman correlations across all bootstrap replicates, which are significantly higher than that of TAJI-M. The bootstrap distribution actually indicates that the Spearman correlation from DIPx is negatively biased, while from TAJI it is slightly positively biased. This means that the gap between the bias-corrected estimates of the Spearman correlations from DIPx and TAJI-M would be even larger; see the Method section for a theoretical explanation.

Performance of DIPx in the AZS dataset. This includes both Test Set 1 (panels a, b, c, f) and Test Set 2 (panels d, e, g). a) Comparison of predicted vs observed synergy scores for all experiments in Test Set 1. b) Comparison of DIPx vs TAJI-M in terms of the correlation between predicted and observed synergy scores from all experiments in Test Set 1. Each boxplot shows the results based on 100 bootstrap replicates of the training set. c) Comparison of DIPx and TAJI-M performance across cell lines in Test Set 1. Each point represents the correlation between the predicted and observed synergy for a given cell line. d) Comparison of DIPx vs TAJI-M in Test Set 2. Each boxplot displays the correlations between the predicted and observed values obtained from 100 bootstrap replicates of the training set. e) Comparison of performance between DIPx and TAJI-M in Test Set 2 in relation to the number of drugs in common (x-axis) between the combinations in the test set and the training set. f) and g) DIPx vs TAJI-M in three groups classified by monotherapy sensivitity of two drugs in a combination in Test Set 1 (f) and Test Set 2 (g).

Furthermore, we compared the performance of DIPx and TAJI-M across all cell lines in Test Set 1 using a Spearman correlation between the predicted and observed synergy scores, as shown in Figure 2c. A majority of the cell lines (63%) are below the diagonal line, indicating that DIPx outperforms TAJI-M in predicting synergy scores for these cell lines.

We also compared the performance of DIPx and TAJI-M in Test Set 2. As expected, the prediction performance of both methods was worse in Test Set 2 than in Test Set 1 since Test Set 2 consists of new combinations absent from the training set. The Spearman correlation of the observed vs predicted synergy using DIPx is 0.26 (95% CI: 0.22–0.30), which is greater than 0.18 (95% CI: 0.16– 0.20) using TAJI-M. Figure 2d show that this result is stable across 100 bootstrap replications of the training set. A similar downward bias for DIPx is observed in the bootstrap distribution.

To investigate the effect of unseen combinations on prediction performance, we divided each combination (drug A + drug B) in Test Set 2 into one of three groups based on the number of individual drugs present in the training set: (i) neither drug A nor drug B in the training set (“no drug”), (ii) either drug A or drug B in the training set (‘one drug’), (iii) and both drugs A and B in the training set (‘two drugs’), as shown in Figure 2e. Overall, both DIPx and TAJI-M showed improved performance as the number of drugs present in the training set increased. For experiments in which both drugs were not in the training set (n = 262), TAJI-M achieved a median correlation of 0.11, while DIPx performed worse with a median correlation of –0.03. For experiments with at least one drug in the training set (n = 2, 499), both methods showed improved performance with median correlations of 0.16 and 0.12 for DIPx and TAJI-M, respectively. When both drugs in an experiment were present in the training set (n = 4, 370), DIPx achieved a median correlation of 0.30, which was better than TAJI-M’s performance (r = 0.22).

Monotherapy drug response profiles have been shown to correlate with synergistic effects and contribute to improving prediction performance, e.g., in TAJI (Li et al., 2018). Here, we compared the performance of DIPx and TAJI-M in relation to monotherapy sensitivity as measured by the IC50 value. We categorized each experiment in the AZS test sets into three groups according to the monotherapy response. Briefly, we first calculated the median sensitivity to monotherapy for each drug A (TA) across all experiments. Measuring the response of a cell line to drug A in an experiment by SA, the drug is considered sensitive if SA ≥ TA. We then compared the synergy of a combination of drugs A and B in relation to the monotherapy sensitivity to both drugs, only one drug, or neither drug.

In Test Set 1, DIPx outperformed TAJI-M in all three groups of monotherapy sensitivity, with the highest performance in the group sensitive to both drugs (median r = 0.48, P-value < 1.79 × 10−27), see Figure 2f. In Test Set 2, TAJI-M performed slightly better in the group with no sensitive drug (median r = 0.21 vs r = 0.20 by DIPx, P-value < 1.26 × 10−5). Interestingly, we found that, while the performance of DIPx improved as the number of monotherapy-sensitive drugs in a combination increased, the performance of TAJI-M decreased, see Figure 2g.

External validation of DIPx in the ONeil dataset

We used a similar computational approach to evaluate the prediction performance of DIPx in relation to the sensitivity of the constituent monotherapies and the impact of unseen combinations in the ONeil dataset. As shown in Figure 3a, the performance of DIPx improved with an increasing number of monotherapy-sensitive drugs in the combination, consistent with the results of Test Set 2 of the AZS data. The highest Spearman correlation between the predicted and observed scores was seen in combinations with two sensitive drugs (median r = 0.11). In relation to the number of drugs in a combination present in the training set, DIPx achieved better performance for combinations with none or one drug in the training set (middle box plot, Figure 3 - figure supplement 1). Poor performance for combinations with two drugs in the training set could be due to the limited number of drug combinations (42/583). We also analyzed the prediction performance of DIPx across the 29 cell lines from 6 different cancer tissues (Figure 3b). Colon cancer (yellow boxplots) and lung cancer cell lines (purple boxplots) showed better validation compared to cell lines from breast, ovarian, melanoma, and prostate cancers.

Prediction performance of DIPx in the ONeil dataset. a) monotherapy sensitivity, b) 29 cell lines from 6 cancer tissues. The y-axis in all box plots shows the Spearman correlation between predicted and observed values in 100 bootstrap replicates.

Figure 3—figure supplement 1. Prediction performance of DIPx in the ONeil dataset, grouped by unseen combinations.

Inference of the mechanism of action based on PAS

The use of PAS in DIPx allows us to infer the potential mechanisms of action of drug combinations while maintaining the prediction performance of the model. For instance, in Test Set 1 of the AZS data, DIPx suggests the involvement of ERBB2 signaling pathways in the Capivasertib + Sapitinib combination, as illustrated by the top pathways depicted in Figure 1e and marked yellow in Figure 4a. This combination therapy has shown promise in overcoming resistance to anti-ERBB2 monotherapy in HER2+ breast cancer (Fujimoto et al., 2020), and ERBB2 has been identified as a key biomarker associated with synergistic responses to this combination in the AZS DREAM Challenge study (Menden et al., 2019).

Inference of pathway importance scores in the AZS dataset. a) Scatter plot showing feature importance (x-axis) vs PAS (y-axis) for the Capivasertib + Sapitinib combination. Pathways with high PAS and feature importance (top 5%) are of particular interest. b, c) Top pathways contributing to the prediction of the combinations in Test Set 1 (b) and Test Set 2 (c). For each pathway, the bar plots show its feature importance. d, f) Functional interaction between the pathway vs driver genes (x-axis) and the pathway vs target genes (y-axis) of the top pathways suggested by DIPx in the SW900 cell line treated with synergistic combination BCL2L1 + AZD5582 (d) and the non-synergistic combination Doxorubicin + AZ12623380 (f). The z-score from network enrichment analysis (NEA) is a measure of functional interaction between two gene sets. A higher z-score indicates a stronger interaction compared to a random permutation of the network. The upper right quadrant (z-score > 1.96) represents pathways that are potentially interesting. e, g) Cartoon illustration of the potential pathways mediated by the synergistic combination of BCL2L1 + AZD5582 (e) and the non-synergistic combination Doxorubicin + AZ12623380 (g).

Figure 4—figure supplement 2. A cartoon illustration of the RAS pathway mediated by the Selumetinib + MK-2206 combination

Figure 4—figure supplement 3. Observed vs predicted inhibition in the SW900 cell line treated by BCL2L1 + AZD5582 and Doxorubicin + AZ12623380 combinations

Figure 4—figure supplement 4. Functional interaction between driver genes, target genes, and top pathways suggested by DIPx in the SW900 cell line treated with BCL2L1 + AZD5582.

Figure 4—figure supplement 5. Functional interaction between driver genes, target genes, and top pathways suggested by DIPx in the SW900 cell line treated with Doxorubicin + AZ12623380.

Figure 4a further shows the distribution of feature importance versus PAS for all pathways for Capivasertib + Sapitinib combination. The feature importance value is calculated using the permutation method of Ishwaran et al. (Ishwaran and Lu, 2019). Our focus is on pathways with high feature importance (e.g., the top 5%) as well as highly activated (top 5% PAS). Therefore, the top-right section of Figure 4a is the interesting region. We present additional examples to further demonstrate the capabilities of DIPx. Figure 4b gives the top pathways of MEDI3622, an ADAM17 inhibitor, in combinations with AKT inhibitors including Capivasertib and MK-2206. These ADAM17 + AKT combinations target multiple parts of the PI3K/AKT pathway through ERBB activation (Menden et al., 2019), which aligns with the potential pathway candidates suggested by DIPx.

One of the key strengths of DIPx is its ability to infer potential mechanisms of both known and novel drug combinations, even in cases where limited biological or clinical information is available. This capability is particularly valuable for new combinations that have not been included in the training set. For instance, in Figure 4c, we present the key pathways identified for the Selumetinib + MK-2206 combination from Test Set 2 of the AZS data. We observe the involvement of RAS signaling, with Selumetinib targeting MEK and MK-2206 targeting AKT, as shown in Figure 4-figure supplement 2. A recent clinical study has used Selumetinib + MK-2206 to target downstream components of the RAS pathway (Chung et al., 2017).

If the drugs in a combination have the same target, the efficacy of the combination is likely similar to that of each individual drug at higher doses, i.e., they will only have an additive effect. So it seems reasonable to hypothesize that a synergistic combination is more likely to occur when the two drugs have different targets (Chen et al., 2015). But how should the targets be related to each other? To investigate this, we examine the pathways suggested by DIPx. First, we choose a synergistic combination of BCL2L1 + AZD5582 in the SW900 cell line for further illustration. The contour plot of the BCL2L1 + AZD5582 inhibition in the SW900 cell line is illustrated in Figure 4 - figure supplement 3a. We first collected the top 15 pathways (ranked by feature importance) for this BCL2L1 + AZD5582 combination suggested by DIPx. The full list of these pathways is shown in Figure 4 - figure supplement 4. Figure 4d illustrates the functional interaction between the genes of the top 15 pathways and the driver genes of the SW900 cell line (x-axis); the target genes of the combination BCL2L1 + AZD5582 (y-axis). To assess the strength of this interaction, we used the network enrichment analysis (NEA) (Alexeyenko et al., 2012), which provides z-score, an enrichment score, indicating the degree of interaction. A higher z-score reflects a stronger interaction between the two gene sets. The top pathways exhibiting high functional interaction with both the driver genes and target genes (z-score > 1.96) are particularly notable, located in the upper right quadrant of Figure 4d. In particular, the apoptosis pathway via NF-kB (highlighted in green) has the highest pathway-target interaction among these pathways. Figure 4e shows the cartoon illustration of the pathway in which the drug BCL2L1 targets BCL-xL and AZD5582 targets XIAP. This suggests an explanation for the observed synergy between the two drugs. Thus, it appears that in this case we get synergy when the drugs target different parts of a driving pathway, either directly or via other functional interactions.

As a negative control, we examine the non-synergistic combination Doxorubicin + AZ12623380, which targets the same gene TOP2; see Figure 4f and g and Figure 4 - figure supplement 3b. We similarly obtain 15 top-ranking pathways according to feature importance, but now we do not expect to see anything obviously relevant to the SW900 cell line (more details in Figure 4 - figure supplement 5). Some pathways that have a high functional interaction with the target genes (upper-left quadrant) have little interaction with the drivers. There are no clearly outlying points in the upper-right quadrant; the two pathways near the boundary are (i) Shen_Smarca2_targets_up, containing genes whose expression negatively correlated with the expression of the SMARCA2 gene in prostate cancer samples, discovered in relation to androgen-induced proliferation in the prostate; and (ii) Kokkinakis_Methione _deprivation_48hr_up, which contains up-regulated genes in melanoma cell-line MEWO cells after 48h of methionine deprivation. They do not appear to be relevant for the lung cancer cell line SW900.

PAS captures the functional interaction of drug targets

As we discuss above, to get synergy, the two drugs in a combination theoretically should not have the same target. However, there is of course no guarantee that two drugs that do not share target genes can produce synergy. In Figure 5a, using the AZS data, we compare the observed drug synergy of combinations of two drugs that share some target genes vs those that do not share any target genes. No significant differences were observed (p-value > 0.72), suggesting that non-overlapping drugs in terms of their targets do not necessarily result in improved drug synergy.

a) Comparison of drug synergy between combinations (drug A + drug B) with vs without overlapping target genes. The numbers in parentheses show the sample sizes of each group. b) Drug synergy between four groups in relation to increased functional interaction between the target genes of the two drugs. c) Comparison between the observed functional interaction (z-score in the network enrichment analysis) and the predicted z-score by PAS.

However, we also observed synergy when the two drugs target different genes in the same pathway. More generally, we hypothesize that synergistic effects occur when the targets have functional interaction. As before, the functional interaction is assessed using NEA (Alexeyenko et al., 2012), where a higher z-score value indicates a stronger functional interaction between the two drugs. Figure 5b shows the observed drug synergy (y-axis) in the AZS data for the four groups defined by the quartile values of the z-scores (x-axis). It indicates that combinations with higher functional interaction are more likely to achieve higher drug synergy, with the highest z-score group (z ∈ (2.97, 29.3]) exhibiting the most favorable drug synergy (median Loewe score = 10.34).

However, when added to the prediction model, the functional interaction z-score did not improve the prediction of synergy (data not shown). Statistically, this can happen if PAS already captures the functional interaction information. To show this, using the AZS training data, we trained a prediction model using PAS as the feature and the functional interaction z-score as the output. We then evaluated the performance of the model in the test set. As shown in Figure 5c, we observed a significant correlation between the predicted and observed z-scores, with a Spearman correlation coefficient of 0.46. This explains why the functional interaction does not give additional predictive power in our model.

Discussion

We have developed and validated DIPx, an advanced computational model that incorporates gene expression and mutation profiles to predict synergistic drug combinations. DIPx performs well against the best performing method in the AstraZeneca-Sanger DREAM Challenge. Through the use of tumor- and patient-specific pathway activation scores, DIPx also provides valuable information on the potential underlying pathways associated with an observed synergistic drug interaction. In addition to rigorous validation using the AZS dataset, DIPx is further validated on the independent ONeil dataset. This comprehensive validation ensures the robustness and reliability of DIPx in predicting drug synergy across different cancers.

The recent availability of large-scale drug combination assay data has allowed the development of realistic prediction models for drug synergy. These datasets offer a substantial number of samples encompassing hundreds of combinations, allowing for extensive validation studies. However, it is important to note that these datasets were generated using different protocols and drug screening techniques. For instance, the AZS data used a 5-by-5 concentration matrix, while the study by ONeil et al. used a 4-by-4 format. In addition, there is limited overlap in the cell lines used among the datasets. These differences pose challenges to the proper validation of prediction methods (Menden et al., 2019).

A particular strength of our study is that we use the best-performing method in the Challenge as a benchmark. This is a convenient but robust benchmarking, as there were 160 teams that participated in the Challenge (73 teams submitted in the final round). Altogether, these teams used practically all of the commonly used machine learning tools; see the summary in Menden et al(Menden et al., 2019). Another strength is our use and validation of the confidence score metric, which captures the statistical uncertainty in the predicted synergy by a single number. This is more convenient for clinical interpretation than the standard prediction interval, because there is a target level for which a combination is considered synergistic, so the score measures our confidence in achieving the target.

Despite promising results, our study has several limitations. First, the use of cell lines as training and validation samples from the AZS and ONeil datasets may not fully capture the heterogeneity present in actual tumors. Second, the computation of PAS relies solely on the primary target genes of the drug combinations, potentially disregarding valuable information from non-primary targets. There could also be off-targets that we do not know about. This limitation might lead to the loss of information about the broader effects of drug combinations. Third, cancer is a heterogeneous disease that occurs in many tissues. Even within a single tissue, cancer exhibits distinct (molecular) subtypes with varying biological mechanisms and clinical outcomes. Since DIPx was developed using pan-cancer datasets, it may not be optimal for tissue-specific predictions.

Last but not least, prediction of previously untrained combinations remains a great challenge. The worst case is for combinations of drugs that were not previously trained, with the Spearman correlation only around 0.1. However, from a clinical perspective, it is perhaps more realistic to look for combinations among drugs previously trained in monotherapy or in other combinations. Improving the prediction for the combination of such drugs would be worthwhile.

Methods and Materials

Pathway activation score for drug combinations

Pathway activation scores (PASs) are the key features in DIPx. The PAS of pathway P in cell line C is calculated for each drug combination (drug A + drug B) and pathway P. Genes in pathway P are grouped into three subgroups: (a) Gu, which includes all the target genes of drugs A and B, as well as the upstream genes of pathway P; (b) Gd , which includes the downstream genes of pathway P; and (c) Gdr, which consists of all the driver genes of cell line C in pathway P. In the example of the ERBB pathway targeted by Capivasertib + Sapitinib. (Figure 1b), Gu consists of ERBB, PI3K, and also AKT; Gd contains MTOR, RAS and MAPK, while Gdr includes TP53 and ERBB2.

The score for upstream activity (PASu) is calculated by the sum of mRNA expression for genes in Gu. Similarly, the scores for the downstream activity (PASd ) and the set of driver genes (PASdr) are calculated from Gd and Gdr. In practice, the genes of the N = 4,762 curated human pathways are provided from the MsigDB database (Liberzon et al., 2015). The target genes of the drugs are collected from the AZS dataset and extended from the DrugBank database (Wishart et al., 2018) and the ChEMBL database (Zdrazil et al., 2024). The extraction of the driver genes of the cell lines is described in the Datasets section.

A pathway based model for drug synergy prediction

The training features of DIPx consist of three components: upstream activity (PASu), downstream activity (PASd ), and driver genes (PASdr), as shown in Figure 1b. The final training matrix has a size of K experiments by 14,286 PASs, where each row corresponds to a specific experiment (drug A + drug B, cell line C).

To address potential sparsity in the training matrix caused by pathways with no target or driver genes, we explored an alternative model with N = 4,762 additional features. Each feature corresponds to a pathway P and is calculated as S(g) * (w1 + w2), where S(g) represents the sum of mRNA expression for all genes in pathway P, and w1 and w2 denote the functional interactions between gene sets: (pathway genes ↔ target genes) and (pathway genes ↔ driver genes), respectively. The functional interactions were estimated using NEA and converted into normal probability scores for w1 and w2. The feature value is zero only when the pathway lacks both targets and driver genes, as well as any interactions with drug targets and driver genes. Additionally, we incorporated the NEA enrichment score between target genes and driver genes into the final matrix. Despite adding these new features, the alternative model did not exhibit any significant improvements in predictive power (data not shown).

For the predictor, we used the random forest algorithm implemented in the randomforestRSC package (with default parameters) in R version 4.0.4. During the development of DIPx, we experimented with various machine learning methods, such as the support vector machine (SVM) and the elastic net. However, we found that these other methods yielded comparable results and that tuning their parameters did not significantly improve prediction performance while requiring extensive additional computations (data not shown). The random forest algorithm in the randomforestRSC package also offers multiple options to calculate the importance of features. In this study, we used the permutation (or Breiman-Cutler) method (Ishwaran and Lu, 2019) to infer the importance of each PAS.

The confidence score (CS) is used to assess the statistical quality of synergy prediction; see Pawitan (2001, Section 5.6) (Pawitan, 2001) for the confidence concept in general. First, as previously defined for example in (Menden et al., 2019), a combination is considered synergistic if the Loewe score is greater than or equal to 20. For each sample s, we have the actual predicted synergy Ps. We then generate Nb = 100 bootstrap replicates of the training data and obtain the bootstrap predictions for the sample: . The CS of Ps is defined as follows:

The bootstrap replicates are also used to evaluate the standard errors (se) of the Spearman correlation between the observed and predicted synergy scores in the test sets. The 95% confidence intervals are computed by the usual formula: se, where is the observed Spearman correlation. Though less frequently used, the bootstrap can also be used for bias correction (Pawitan, 2001, Section 5.2). Bias occurs if there is a nontrivial gap between the observed estimate and the mean of the bootstrap replications. Theoretically, where F is the underlying data distribution. So, the bias-corrected estimate should be

In practice, the bias is estimated by where are the bootstrap replicates of . When the estimated bias is negative, as we observed for DIPx, the bias-corrected estimate is shifted upward. And vice versa, if the bias is positive, as observed for TAJI-M, the corrected estimate is shifted downward.

Datasets

AstraZeneca-Sanger (AZS) DREAM Challenge dataset

The AZS DREAM Challenge is a rigorous competition in the effort to systematically develop and validate drug synergy prediction methods. Indicating the strong interest in the topic, 160 international teams (Menden et al., 2019) participated in the Challenge. It was organized into two subchallenges: i) Prediction for known (tested) combination and ii) Prediction for unknown (untested) drug combinations. The final dataset comprised 11,576 experiments from 85 cell lines and 910 combinations. The gene expression data of these cell lines was obtained from Affymetrix microarray ( Menden et al., 2019). H owever, to ensure consistency between the AZS dataset and the Oneil dataset (O’Neil et al., 2016) (which did not provide gene expression profiles of cell lines), we utilized gene expression data from the Cancer Cell Line Encyclopedia (CCLE) cohort (Ghandi et al., 2019).

Out of the 85 cell lines, we identified 75 cell lines with available gene expression data in the CCLE cohort, resulting in a total of 10,154 experiments involving 910 combinations used in our study. Supplementary File S1 shows the list of 75 cell lines. For the validation of the prediction model, the data were split into a training set (n = 2060) and two test sets (n = 963 and 7131) according to subchallenges 1 and 2, respectively. The first test set contains experiments from 167 combinations (of 69 single drugs) that are also in the training set. The second test set includes experiments with 729 drug combinations that are not in the training set.

We collected gene expression data of 75 cell lines, measuring the transcript per million (TPM) of 37,222 genes, of the CCLE cohort downloaded from the DepMap Portal (Tsherniak et al., 2017). The gene expression data was logarithmically transformed to the base 2 scale for downstream analysis. Additionally, we obtained potential driver genes for these cell lines, including both mutations and fusion genes, from the DepMap Portal. The portal provides information on mutations in 1,637 protein-coding genes associated with cancer biology in a collection of 1,030 cell lines.

To filter the list of mutations, we focused on those occurring in at least 2.5% of the total cell lines. Subsequently, we extracted the list of mutations specific to the 75 cell lines under investigation. For fusion genes, we focused on those present in the Miltelman database (Mitelman, 2022) and occurring at least twice, considering them as relevant for our analysis. The final list of potential driver genes for the 75 cell lines can be found in Supplementary File S1. On average, each cell line had a median of 29 potential driver genes.

For the drug synergy data, we used a 5-by-5 concentration matrix provided by the Challenge. Drug synergy values were estimated using the Loewe reference model from Combenefit (Di Veroli et al., 2016).

ONeil dataset

ONeil dataset is a large-scale drug synergy screening dataset from Merck&Co company (O’Neil et al., 2016). A total of 23,062 experiments with 583 unique drug combinations (38 monotherapy drugs) was carried out on 38 cancer cell lines by a 4-by-4 drug concentration matrix. Out of 38 cell lines, we found 29 cell lines with available gene expression data from the DepMap Portal. The detail of 29 cell lines is described in Supplementary File S1. The gene expression data of 37,222 genes from 29 cell lines, as well as the driver genes of these cell lines, were collected from the DepMap Portal using the same procedure as in the AZS dataset. The original release of this dataset provides only the raw data on drug synergy. Here, we calculated the Loewe synergy score for each experiment using Combenefit (Di Veroli et al., 2016). In total, we obtained 16,907 experiments for 583 combinations in 29 cell lines for further analysis. Drug targets of 38 monotherapy drugs were collected from the DrugBank database (Wishart et al., 2018) and the ChEMBL database (Zdrazil et al., 2024).

eLife Assessment

This study presents a valuable report on a machine-learning tool for predicting synergistic drug combinations for cancer treatment. However, the evidence supporting the claims of the authors is incomplete, as the reported model shows some evidence of overfitting, and the claims of the authors could be strengthened if additional validation experiments were performed. The work will be of interest to oncologists and medical biologists working on cancer.

Reviewer #1 (Public Review):

The authors introduce DIPx, a deep learning framework for predicting synergistic drug combinations for cancer treatment using the AstraZeneca-Sanger (AZS) DREAM Challenge dataset. While the approach is innovative, I have the following concerns and comments which hopefully will improve the study's rigor and applicability, making it a more powerful tool in the real clinical world.

(1) Test Set 1 comprises combinations already present in the training set, likely leading overfitting issue. The model might show inflated performance metrics on this test set due to prior exposure to these combinations, not accurately reflecting its true predictive power on unknown data, which is crucial for discovering new drug synergies. The testing approach reduces the generalizability of the model's findings to new, untested scenarios.

(2) The model struggles with predicting synergies for drug combinations not included in its training data (showing only a Spearman correlation of 0.26 in Test Set 2). This limits its potential for discovering new therapeutic strategies. Utilizing techniques such as transfer learning or expanding the training dataset to encompass a wider range of drug pairs could help to address this issue.

(3) The use of pan-cancer datasets, while offering broad applicability, may not be optimal for specific cancer subtypes with distinct biological mechanisms. Developing subtype-specific models or adjusting the current model to account for these differences could improve prediction accuracy for individual cancer types.

(4) Line 127, "Since DIPx uses only molecular data, to make a fair comparison, we trained TAJI using only molecular features and referred to it as TAJI-M.". TAJI was designed to use both monotherapy drug-response and molecular data, and likely won't be able to reach maximum potential if removing monotherapy drug-response from the training model. It would be critical to use the same training datasets and then compare the performances. From Figure 6 of TAJI's paper (Li et al., 2018, PMID: 30054332) , i.e., the mean Pearson correlation for breast cancer and lung cancer is around 0.5 - 0.6.

The following 2 concerns have been included in the Discussion section which is great:

(1) Training and validating the model using cell lines may not fully capture the heterogeneity and complexity of in vivo tumors. To increase clinical relevance, it would be beneficial to validate the model using primary tumor samples or patient-derived xenografts.

(2) The Pathway Activation Score (PAS) is derived exclusively from primary target genes, potentially overlooking critical interactions involving non-primary targets. Including these secondary effects could enhance the model's predictive accuracy and comprehensiveness.

Reviewer #2 (Public Review):

Trac, Huang, et al used the AZ Drug Combination Prediction DREAM challenge data to make a new random forest-based model for drug synergy. They make comparisons to the winning method and also show that their model has some predictive capacity for a completely different dataset. They highlight the ability of the model to be interpretable in terms of pathway and target interactions for synergistic effects. While the authors address an important question, more rigor is required to understand the full behavior of the model.

Major Points

(1) The authors compare DIPx to the winning method of the DREAm challenge, TAJI to show that from molecular features alone they retrain TAJI to create TAJI-M without the monotherapy data inputs. They mention that "of course, we could also use such data in DIPx...", but they never show the behaviour of DIPx with these data. The authors need to demonstrate that this statement holds true or else compare it to the full TAJI.

(2) It would be neat to see how the DIPx feature importance changes with monotherapy input. For most realistic scenarios in which these models are used robust monotherapy data do exist.

(3) In Figure 2, the authors compare DIPx and TAJI-M on various test sets. If I understood correctly, they also bootstrapped the training set with n=100 and reported all the model variants in many of the comparisons. While this is a nice way of showing model robustness, calculating p-values with bootstrapped data does not make sense in my opinion as by increasing the value of n, one can make the p-value arbitrarily small. The p-value should only be reported for the original models.

(4) From Figures 2 and 3, it appears DIPx is overfit on the training set with large gaps in Spearman correlations between Test Set 2/ONeil set and Test Set 1. It also features much better in cases where it has seen both compounds. Could the authors also compare TAJI on the ONeil dataset to show if it is as much overfit?

Reviewer #3 (Public Review):

Summary:

Predicting how two different drugs act together by looking at their specific gene targets and pathways is crucial for understanding the biological significance of drug combinations. Such combinations of drugs can lead to synergistic effects that enhance drug efficacy and decrease resistance. This study incorporates drug-specific pathway activation scores (PASs) to estimate synergy scores as one of the key advancements for synergy prediction. The new algorithm, Drug synergy Interaction Prediction (DIPx), developed in this study, uses gene expression, mutation profiles, and drug synergy data to train the model and predict synergy between two drugs and suggests the best combinations based on their functional relevance on the mechanism of action. Comprehensive validations using two different datasets and comparing them with another best-performing algorithm highlight the potential of its capabilities and broader applications. However, the study would benefit from including experimental validation of some predicted drug combinations to enhance its reliability.

Strengths:

The DIPx algorithm demonstrates the strengths listed below in its approach for personalized drug synergy prediction. One of its strengths lies in its utilization of biologically motivated cancer-specific (driver genes-based) and drug-specific (target genes-based) pathway activation scores (PASs) to predict drug synergy. This approach integrates gene expression, mutation profiles, and drug synergy data to capture information about the functional interactions between drug targets, thereby providing a potential biological explanation for the synergistic effects of combined drugs. Additionally, DIPx's performance was tested using the AstraZeneca-Sanger (AZS) DREAM Challenge dataset, especially in Test Set 1, where the Spearman correlation coefficient between predicted and observed drug synergy was 0.50 (95% CI: 0.47-0.53). This demonstrates the algorithm's effectiveness in handling combinations already in the training set. Furthermore, DIPx's ability to handle novel combinations, as evidenced by its performance in Test Set 2, indicates its potential for extrapolating predictions to new and untested drug combinations. This suggests that the algorithm can adapt to and make accurate predictions for previously unencountered combinations, which is crucial for its practical application in personalized medicine. Overall, DIPx's integration of pathway activation scores and its performance in predicting drug synergy for known and novel combinations underscore its potential as a valuable tool for personalized prediction of drug synergy and exploration of activated pathways related to the effects of combined drugs.

Weaknesses:

While the DIPx algorithm shows promise in predicting drug synergy based on pathway activation scores, it's essential to consider its limitations. One limitation is that the algorithm's performance was less accurate when predicting drug synergy for combinations absent from the training set. This suggests that its predictive capability may be influenced by the availability of training data for specific drug combinations. Additionally, further testing and validation across different datasets (more than the current two datasets) would be necessary to assess the algorithm's generalizability and robustness fully. It's also important to consider potential biases in the training data and ensure that DIPx predictions are validated through empirical studies including experimental testing of predicted combinations. Despite these limitations, DIPx represents a valuable step towards personalized prediction of drug synergy and warrants continued investigation and improvement. It would benefit if the algorithm's limitations are described with some examples and suggest future advancement steps.

Author response:

The following is the authors’ response to the original reviews.

Reviewer #1 (Public Review):

The authors introduce DIPx, a deep learning framework for predicting synergistic drug combinations for cancer treatment using the AstraZeneca-Sanger (AZS) DREAM Challenge dataset. While the approach is innovative, I have the following concerns and comments which hopefully will improve the study's rigor and applicability, making it a more powerful tool in the real clinical world.

We thank to the reviewer for recognizing the innovative aspects of DIPx and for sharing their valuable comments to further refine and strengthen our study. Those comments are carefully addressed in the following point-by-point response.

(1) Test Set 1 comprises combinations already present in the training set, likely leading overfitting issue. The model might show inflated performance metrics on this test set due to prior exposure to these combinations, not accurately reflecting its true predictive power on unknown data, which is crucial for discovering new drug synergies. The testing approach reduces the generalizability of the model's findings to new, untested scenarios.

From a clinical perspective, it is useful to test whether a known (previously tested) combination can work for a new patient, which is the purpose of Test Set 1. There is no danger overfitting here, because the test set is completely independent of the discovery set, so had we only discovered a false positive the test set would not have more than power than expected under the null. Predicting the effectiveness of unknown drug combinations (Test Set 2) is indeed an important and more challenging goal of synergy prediction, but it is statistically a distinct problem. The two test sets were previously designed by the AZS DREAM Challenge [PMID: 31209238].

We have performed cross-validation on the dataset and demonstrated that the result of DIPx for Test Set 1 is not overfitting. Indeed, Figure 2—figure supplement 1 shows the 10-fold cross validation results for the training set. The median Spearman correlation between the predicted and observed Loewe scores across the 10 folds of cross-validation is 0.48, which is close to the correlation of 0.50 in Test Set 1 (red star). We have added the cross-validation results to the “Validation and Comparisons in the AZS Dataset” section (page 4).

(2) The model struggles with predicting synergies for drug combinations not included in its training data (showing only a Spearman correlation of 0.26 in Test Set 2). This limits its potential for discovering new therapeutic strategies. Utilizing techniques such as transfer learning or expanding the training dataset to encompass a wider range of drug pairs could help to address this issue.

We agree that this is an important limitation for the discovery of new therapeutic strategies. While transfer learning or expanding the training dataset could indeed help address this issue, implementing these approaches would require access to more comprehensive data, which is currently limited due to the scarcity of drug combination datasets. As more drug combination data become available in future, we plan to expand the training set to better cover a wider range of drug combinations and apply the transfer learning method to improve prediction accuracy. We have added a discussion on this in the Discussion Section.

(3) The use of pan-cancer datasets, while offering broad applicability, may not be optimal for specific cancer subtypes with distinct biological mechanisms. Developing subtype-specific models or adjusting the current model to account for these differences could improve prediction accuracy for individual cancer types.

We agree with the reviewer that the current settings of DIPx might not be optimal for specific cancers due to the cancer heterogeneity. However, building subtype-specific models is currently constrained by limitation of data availability, which in turn restricts their predictive power. In the Discussion section, we mention this as one of DIPx's limitations and suggest future improvements in cancer-specific models.

(4) Line 127, "Since DIPx uses only molecular data, to make a fair comparison, we trained TAJI using only molecular features and referred to it as TAJI-M.". TAJI was designed to use both monotherapy drug-response and molecular data, and likely won't be able to reach maximum potential if removing monotherapy drug-response from the training model. It would be critical to use the same training datasets and then compare the performances. From Figure 6 of TAJI's paper (Li et al., 2018, PMID: 30054332) , i.e., the mean Pearson correlation for breast cancer and lung cancer is around 0.5 - 0.6.

It is true that using monotherapy drug responses can enhance the performance of TAIJI as described in its original paper. In fact, TAIJI builds separate prediction modules for molecular data and monotherapy drug-response data, then combine their results to obtain the final prediction. In our paper we prioritize the exploration of molecular mechanisms in drug combinations while achieving performance comparable to the molecular model of TAIJI. DIPx can be expected to achieve similarly improved performance if we integrate the monotherapy drug response data using the same approach.

My major concerns were listed in the public review. Here are some writing issues:

(5) Some content in the Results section looks like a discussion: i.e, L129, "The extra information from the use of monotherapy data in TAJI is rather small, approximately 10% increase in the overall Spearman correlation, and, of course, we could also use such data in DIPx, so it is more convenient and informative to focus the comparisons on prediction based on molecular data alone."; L257, "As we discuss above, to get synergy, the two drugs in a combination theoretically should not have the same target. However, there is of course no guarantee that two drugs that do not share target genes can produce synergy. ".

We have revised the texts and moved them to the Discussion section.

Reviewer #2 (Public Review):

Trac, Huang, et al used the AZ Drug Combination Prediction DREAM challenge data to make a new random forest-based model for drug synergy. They make comparisons to the winning method and also show that their model has some predictive capacity for a completely different dataset. They highlight the ability of the model to be interpretable in terms of pathway and target interactions for synergistic effects. While the authors address an important question, more rigor is required to understand the full behavior of the model.

We thank the reviewer for his/her time and effort in carefully reading the manuscript and acknowledging the significance of the study.

Major Points

(1) The authors compare DIPx to the winning method of the DREAm challenge, TAJI to show that from molecular features alone they retrain TAJI to create TAJI-M without the monotherapy data inputs. They mention that "of course, we could also use such data in DIPx...", but they never show the behaviour of DIPx with these data. The authors need to demonstrate that this statement holds true or else compare it to the full TAJI.

This is similar to point 4 raised by Reviewer 1 regarding the exclusive use of molecular data in DIPx. In fact, TAIJI uses separate prediction modules for molecular data and drugresponse data which are then combined to obtain the final results. While integrating monotherapy drug data could enhance DIPx’s overall performance, for example, simply replacing TAIJI’s molecular model with DIPx in the full TAIJI to achieve comparable results, this is not the primary goal of DIPx. Our focus is on exploring the potential molecular mechanisms of drug action. Using only molecular data allows for more convenient and intuitive inference of pathway importance compared to integrating multiple data types.

We have revised the related text with the discussion in section “Validation and comparisons in the AZS dataset” of the main text.

(2) It would be neat to see how the DIPx feature importance changes with monotherapy input. For most realistic scenarios in which these models are used robust monotherapy data do exist.

Indeed, some existing models incorporate monotherapy data into their predictions; for example, a recent study [PMID: 33203866] uses only monotherapy data to predict drug combinations. TAIJI, as discussed in Point 1, uses separate models for monotherapy and molecular data. In general, both data types can be integrated into a single prediction model, allowing for the consideration of feature importance from both. While such an approach can highlight features contributing to predictive performance, the significance of a monotherapy feature does not necessarily indicate the activated pathways of a synergistic drug combination, which is the primary focus of our study. For this reason, we have excluded monotherapy data from DIPx.

(3) In Figure 2, the authors compare DIPx and TAJI-M on various test sets. If I understood correctly, they also bootstrapped the training set with n=100 and reported all the model variants in many of the comparisons. While this is a nice way of showing model robustness, calculating p-values with bootstrapped data does not make sense in my opinion as by increasing the value of n, one can make the p-value arbitrarily small.

The p-value should only be reported for the original models.

The reviewer is correct that we cannot compute the p-value by using an independent twosample test, because the bootstrap correlation values are based on the same data. However, p-values can still be computed to compare the two prediction models using the bootstrap. Theoretically, the bootstrap can be used to compute a confidence interval for the differential correlation in the test set. However, there is a close relationship between p-values and confidence intervals (see Pawitan, 2001, chapter 5; particularly p.134). Specifically, in this case, we compute the p-value as follows: (1) For each bootstrap, (i) compute the Spearman correlation between the predicted and observed scores in the test set for DIPx and TAIJI-M.

Denote this by r1 and r2. (ii) compute the difference in the Spearman correlations d= (r1-r2). (2). Repeat the bootstrap n=100 times. (3). Compute the minimum of these two proportions:

proportion of d<0 or proportion of d>0. (4). The two-sided p-value = 2x the minimum proportion in (3). To overcome the limited bootstrap sample size, we use the normal approximation in computing the proportions in (3). Note that in this method of computing the p-value, larger numbers of bootstrap replicates do not produce more significant results.

We have re-computed the p-values using this method and added this text to the ‘Methods and Materials’ Section.

(4) From Figures 2 and 3, it appears DIPx is overfit on the training set with large gaps in Spearman correlations between Test Set 2/ONeil set and Test Set 1. It also features much better in cases where it has seen both compounds. Could the authors also compare TAJI on the ONeil dataset to show if it is as much overfit?

The poor performance in ONeil dataset is not due to overfitting as such, but more likely due structural differences between the training and ONeil datasets. (To investigate the overfitting issue, we have conducted a 10-fold cross validation in the AZS training set. The median correlation between the predicted and observed Loewe score across ten folds is 0.48, which is comparable to the median of 0.50 in the Test Set 1. Therefore, the model does not suffer from overfitting issue. We have added this cross-validation result in the Section “Validation and Comparisons in the AZS Dataset” (page 4)).

We have now obtained TAIJI’s results on the ONeil dataset. TAIJI-M relies on a gene-gene interaction network to integrate the indirect drug targeting effects. This approach limits its applicability to new datasets, as it can only predict synergy scores for drug combinations present in the training dataset. Among the set of drug combinations present in the training set (n = 1102), both DIPx and TAIJI-M perform poorly, with Spearman correlations between predicted and observed synergy scores of 0.09 and 0.05, respectively.

(Additional note: The original version of TAIJI-M uses gene expression, CNV, mutation, and methylation data. However, there is no methylation data in the ONeil dataset, so we retrained TAIJI-M without the methylation features. According to the final report of TAIJI in the challenge (https://www.synapse.org/Synapse:syn5614689/wiki/396206), Guan et al. reported that methylation features do not contribute to prediction performance in the postchallenge analysis. This means that retraining TAIJI-M without the methylation data will not materially affect the comparison between DIPx and TAIJI-M on the ONeil dataset.)

Minor Points:

(5) Pg 4, line 130: Citation needed for 10% contribution of monotherapy.

(6) The general language of this paper is informal at times. I request the authors to refine it a bit.

We thank the reviewer for pointing this out. We have added the appropriate citation for the statement and carefully revised the text to make it more formal.

Reviewer #3 (Public Review):

Summary:

Predicting how two different drugs act together by looking at their specific gene targets and pathways is crucial for understanding the biological significance of drug combinations. Such combinations of drugs can lead to synergistic effects that enhance drug efficacy and decrease resistance. This study incorporates drug-specific pathway activation scores (PASs) to estimate synergy scores as one of the key advancements for synergy prediction. The new algorithm, Drug synergy Interaction Prediction (DIPx), developed in this study, uses gene expression, mutation profiles, and drug synergy data to train the model and predict synergy between two drugs and suggests the best combinations based on their functional relevance on the mechanism of action. Comprehensive validations using two different datasets and comparing them with another best-performing algorithm highlight the potential of its capabilities and broader applications. However, the study would benefit from including experimental validation of some predicted drug combinations to enhance its reliability.

Strengths:

The DIPx algorithm demonstrates the strengths listed below in its approach for personalized drug synergy prediction. One of its strengths lies in its utilization of biologically motivated cancer-specific (driver genes-based) and drug-specific (target genes-based) pathway activation scores (PASs) to predict drug synergy. This approach integrates gene expression, mutation profiles, and drug synergy data to capture information about the functional interactions between drug targets, thereby providing a potential biological explanation for the synergistic effects of combined drugs. Additionally, DIPx's performance was tested using the AstraZeneca-Sanger (AZS) DREAM Challenge dataset, especially in Test Set 1, where the Spearman correlation coefficient between predicted and observed drug synergy was 0.50 (95% CI: 0.470.53). This demonstrates the algorithm's effectiveness in handling combinations already in the training set. Furthermore, DIPx's ability to handle novel combinations, as evidenced by its performance in Test Set 2, indicates its potential for extrapolating predictions to new and untested drug combinations. This suggests that the algorithm can adapt to and make accurate predictions for previously unencountered combinations, which is crucial for its practical application in personalized medicine. Overall, DIPx's integration of pathway activation scores and its performance in predicting drug synergy for known and novel combinations underscore its potential as a valuable tool for personalized prediction of drug synergy and exploration of activated pathways related to the effects of combined drugs.

Weaknesses:

While the DIPx algorithm shows promise in predicting drug synergy based on pathway activation scores, it's essential to consider its limitations. One limitation is that the algorithm's performance was less accurate when predicting drug synergy for combinations absent from the training set. This suggests that its predictive capability may be influenced by the availability of training data for specific drug combinations. Additionally, further testing and validation across different datasets (more than the current two datasets) would be necessary to assess the algorithm's generalizability and robustness fully. It's also important to consider potential biases in the training data and ensure that DIPx predictions are validated through empirical studies including experimental testing of predicted combinations. Despite these limitations, DIPx represents a valuable step towards personalized prediction of drug synergy and warrants continued investigation and improvement. It would benefit if the algorithm's limitations are described with some examples and suggest future advancement steps.

We are grateful to the reviewer for the thoughtful and encouraging comments, and for the time and effort to read our manuscript. We have carefully addressed them in our revision.

Reviewer #3 (Recommendations For The Authors):

The authors could consider some of the recommendations below to further improve the DIPx algorithm and its application in personalized drug synergy prediction. Firstly, expanding the training dataset to include a broader range of drug combinations could improve the algorithm's predictive capabilities, especially for novel combinations. This would help address the observed decrease in performance when predicting drug synergy for combinations absent from the training set. This could help assess the robustness of the algorithm and provide a more comprehensive evaluation of its performance for untrained combinations to strengthen its application.

We agree that expanding the training dataset with a broader range of drug combinations would likely improve performance. However, the vast number of possible combinations, along with the associated cost of the experiment, limits the availability of drug combination data. To increase the size of the training data, we could combine different studies, but data from different studies are often generated using different protocols and experimental settings, introducing biases that complicate the integration. As technology continues to advance, we anticipate that more standardized and comprehensive data will become available in the future, which will help address this issue.

Furthermore, the authors may consider incorporating additional features or data sources, such as drug-specific characteristics, i.e., availability of the drug, to enrich the information utilized by the algorithm. This could potentially improve the accuracy of the predictions and provide a more holistic understanding of the factors contributing to drug synergy.

Indeed, incorporating additional information such as monotherapy data and drug-specific characteristics, as in TAIJI’s approach, could enhance overall prediction performance. As discussed in Point 5 below, the current study is focused on exploring the potential molecular mechanisms of drug combinations, rather than optimizing overall prediction accuracy. However, in its application, it is natural to add the monotherapy or drug-specific information into the algorithm, as done in TAIJI.

Finally, conducting experimental studies to validate the predictions generated by DIPx in laboratory-based cell lines would be essential to confirm its accuracy and reliability. This could involve a few drug IC50 experimental validations of predicted synergistic drug combinations and their associated pathway activations to strengthen the algorithm's clinical relevance. By considering these recommendations, the authors can further refine and advance the DIPx algorithm.

We agree that laboratory-based validation, such as IC50 experiments for predicted synergistic drug combinations and pathway activations, would indeed strengthen the clinical relevance of the algorithm. We hope future studies can build on this work by incorporating this experimental validation.

Below are my specific comments:

Major comments:

(1) The description of all the outputs of the DIPX algorithm is not clearly explained. It is unclear whether it provides only the Loewe score, the confidence score, the PAS score, or all of them. It is necessary to clarify the output of the proposed algorithm to guide the reader on what to expect while using it. The steps from PASs to synergy scores are not well explained.

We apologize for the lack of clarity. Regarding the outputs of DIPx, for any triplet (drug A + drug B, cell line C), DIPx provides both the predicted Loewe score and the corresponding confidence score as the output. PASs are used as the input data for the random forest algorithm, which processes PASs into the synergy score. We do not provide the details in the manuscript, but refer to the article by Ishwaran H et al., (2021). We have revised the first paragraph of the 'A Pathway-Based Drug Synergy Prediction Model' section (page 3) and Figure 1 to improve the presentation of the method.

(2) In Figure 1, the predicted Loewe score for the Capivasertib + Sapitinib combination is not provided. However, Figures 1e and 4a show the pathways with the highest contribution for this combination. What is the predicted Loewe score for the Capivasertib + Sapitinib combination?

Figures 1e and 4a presents the pathways with the highest contribution for the combination which are identified based on the drug-combination data from 12 cell lines, not a single data point.

We have added the median Loewe score (=7.6) across 12 cell lines in the test sets (Test 1 + Test 2) for the Capivasertib + Sapitinib combination in Figure 1e and reported related information for this combination in Supplementary Table S1. Additionally, we revised the 'Inference of the Mechanism of Action Based on PAS' section (page 7) to clarify the pathway importance inference.

(3) In Figure 1d, the combination of doxorubicin + AZ12623380 is predicted to exhibit high Loewe synergy, with a confidence score of 0.33. It is important to provide details of this prediction, including the pathway predictions, and to explain why the model suggested high synergy. Although Figure 4f contains information, it seems to be listed for the observed Loewe score rather than the predicted score provided in Figure 1d. DIPx predicts the doxorubicin + AZ12623380 combination to be synergistic, while in Figure 4, it is labeled as a non-synergistic combination. It is necessary for the authors to clearly indicate which illustration represents the predicted outcome and which hypothesis is based on the observed Loewe score.

In Figure 1d, we reported both predicted and observed Loewe score for the experiment (combination = doxorubicin + AZ12623380, cell line = SW900). Although the predicted score is high, a confidence score of 0.33 indicates that there is a low chance of the prediction is synergistic. And this is indeed confirmed by the non-synergistic observed score of -6, so it does not merit further investigation. This example highlights the value of the confidence score to supplement the predicted values.

(4) Figure 3 - The external validation using ONeil requires more rigorous analysis to understand the biological significance of the predictions. It is important to provide pathway activation scores and their potential mechanism of action predicted by the DIPx algorithm when working with a new dataset. Additionally, including the predictions of TAIJI-M on the ONeil dataset would be beneficial for comparing the performance of both algorithms on a new dataset.

We have included an example of potential pathways related to the MK2206 + Erlotinib combination in the ONeil cohort, as inferred by DIPx, in the last paragraph of the 'Inference of the Mechanism of Action Based on PAS' section (page 9). In this example, we identify 'Metabolism by CYP Enzymes' as the most significant pathway associated with this combination, which aligns with previous studies that both MK2206 and Erlotinib are metabolized by the CYP enzyme families [PMID: 24387695].

Regarding the prediction of TAIJI-M on the ONeil dataset, we have a similar request in question 4 from Reviewer 2, which we have carefully addressed above. Briefly, due to differences between two datasets, we retrained TAIJI-M without methylation data to enable prediction on the ONeil dataset. (As previously reported, methylation data did not significantly contribute to the results of TAIJI, and TAIJI-M can only predict synergy scores for drug combinations present in the training set.) Focusing on this subset of drug combinations, both TAIJI-M and DIPx perform poorly, with Spearman correlations of r=0.05 and r=0.09, respectively. The poor performance could be attributed to the limited overlap of drugs between the ONeil dataset and the AZS DREAM Challenge dataset.

(5) TAIJI by Li et al., 2018 reported a high prediction correlation (0.53) in their study, while the modified version of TAIJI, TAJI-M, shows a lower prediction correlation in this study. The authors should clarify why the performance decreased when using the same dataset. Is it because only molecular data was used, excluding the monotherapy drug-response data? There is a spelling error in calling the algorithm - it is reported as TAIJI by Li et al., 2018, whereas this study calls it TAJI - an "I" is missing in TAIJI throughout the manuscript.

Indeed, TAIJI-M has a lower prediction correlation (0.38) compared to the full TAIJI model (0.53), which includes the monotherapy data. Some studies such as [PMID: 33203866] even use only monotherapy data in prediction of drug combinations, suggesting the importance of monotherapy data in the drug-combination prediction. However, DIPx focuses on exploration of potential molecular mechanisms of drug combinations rather than overall prediction results, therefore, we exclude the monotherapy data from analysis. We have discussed on this in the 'Validation and Comparisons in the AZS Dataset' section (page 4).

We thank the reviewer for pointing the spelling error for TAIJI; this has been corrected throughout the manuscript.

(6) The authors should provide the predicted versus observed Loewe scores for all the combinations as a supplementary file. This would benefit the readers who want to replicate the results in the future. In the same way, including a sample output for the toy dataset on GitHub is required to assess the performance of the DIPx algorithm by a new user.

All predicted and observed drug synergy scores are given in Supplementary Table S2. We also have already uploaded a simple example on our GitHub page, along with detailed instructions for users on how to run the method, including generating PAS and training the prediction model. Since we do not have permission to host data from the AZS DREAM Challenge and the ONeil datasets on our GitHub page, users can download these datasets separately and directly apply the provided code.

(7) GitHub can include all the input and output data to reproduce the correlation plots in the manuscript. GitHub could also include the modified version of TAIJI-M and its corresponding input for comparison. The methods section should include how TAIJI was performed.

We have uploaded all the codes and related data to the GitHub page to allow replication of all correlation plots in the manuscript. TAIJI-M represents the molecular model of the full TAIJI model. Both TAIJI-M and TAIJI are documented on the GitHub page of the original study. We have also included a link to the source code for TAIJI-M and TAIJI in the 'Data Availability' section.

(8) Figure 5 - the data associated with this figure needs to be provided as supplementary listing the predicted values of Loewe scores for all the combinations.

We report the associated data including the median of predicted and observed Loewe scores related to Figure 5c in Supplementary Table S2.

Minor comments:

(9) Abbreviations for the pathways are not included.

We have included a list of abbreviations for all relevant pathways in Supplementary Table S5.

(10) Line: 369. What is considered as bias correction? This needs to be explained.

Bias correction refers to adjusting the original estimate of the Spearman correlation between the predicted and observed Loewe scores when there is a systematic difference between the estimates obtained from the bootstrap samples and the original correlation estimate. We revised the related text in page 13 to improve the explanation.

(11) Line 364. Formulae or details for calculating actual predicted synergy (Ps) are missing.

The predicted Loewe score, Ps, is the output of the regression random forest model. For simplicity, we do not describe the details in the manuscript, but refer to the description of the method article (Ishwaran H et al., 2021). We have revised the text accordingly.

Version 2

Abstract

Abstract

Targeted monotherapies for cancer often fail due to inherent or acquired drug resistance. By aiming at multiple targets simultaneously, drug combinations can produce synergistic interactions that increase drug effectiveness and reduce resistance. Computational models based on the integration of omics data have been used to identify synergistic combinations, but predicting drug synergy remains a challenge. Here, we introduce DIPx, an algorithm for personalized prediction of drug synergy based on biologically motivated tumor- and drug-specific pathway activation scores (PASs). We trained and validated DIPx in the AstraZeneca-Sanger (AZS) DREAM Challenge dataset using two separate test sets: Test Set 1 comprised the combinations already present in the training set, while Test Set 2 contained combinations absent from the training set, thus indicating the model”s ability to handle novel combinations. The Spearman correlation coefficients between predicted and observed drug synergy were 0.50 (95% CI: 0.47–0.53) in Test Set 1 and 0.26 (95% CI: 0.22–0.30) in Test Set 2, compared to 0.38 (95% CI: 0.34–0.42) and 0.18 (95% CI: 0.16–0.20), respectively, for the best performing method in the Challenge. We show evidence that higher synergy is associated with higher functional interaction between the drug targets, and this functional interaction information is captured by PAS. We illustrate the use of PAS to provide a potential biological explanation in terms of activated pathways that mediate the synergistic effects of combined drugs. In summary, DIPx can be a useful tool for personalized prediction of drug synergy and exploration of activated pathways related to the effects of combined drugs.

Introduction

Targeted therapies such as specific inhibitors are the most promising class of cancer drugs, but often fail or achieve only temporary remission due to inherent or acquired resistance. Theoretically, by aiming at multiple targets simultaneously, drug combinations can produce a synergistic interaction that increases drug effectiveness and reduces resistance and the chances of relapse (Medicine, 2017; Pemovska et al., 2018; Plana et al., 2022). This is illustrated in the combination of a BRAF inhibitor dabrafenib with a MEK inhibitor trametinib, which suppresses paradoxical reactivation and resistance observed in patients with BRAF-mutated melanoma treated with dabrafenib alone (Zhong et al., 2022; Banzi et al., 2016). This recently approved combination has been shown to improve progression-free and overall survival rates (Subbiah et al., 2023).

The discovery of effective drug combinations has traditionally relied on expert knowledge and understanding of known biological mechanisms (Li et al., 2015). However, this expert-based approach has limited scope to come up with novel combinations. Furthermore, ideally, novel combinations are clinically tested, but it is practically impossible to test all reasonable combinations in a clinical setting. Computational models of drug synergy have shown some potential for personalized prediction of synergistic combinations (Güvenç Paltun et al., 2021; Wu et al., 2022; Kong et al., 2022). These models are typically based on the integration of patient-specific molecular data, such as mutation profiles, gene expression, and drug response information (Güvenç Paltun et al., 2021). For example, TAIJI, the best performing method in the AstraZeneca-Sanger DREAM Challenge, uses these diverse data types to predict drug synergy (Li et al., 2018). The drug combinations predicted to be effective will expand the therapeutic options while maintaining the same level of adverse effects profile. However, despite the advantages offered by modern machine learning methodologies and the availability of large-scale datasets, prediction of synergistic combinations and validating computational models remains challenging. For example, drug screening protocols often vary across studies, and there is a limited overlap in tested drugs and cell lines, complicating the external validation of these models. Additionally, the reliance on ‘black-box’ machine learning approaches hinders the exploration of underlying molecular mechanisms driving synergistic combinations.

To address this limitation, several studies have introduced statistical and computational ap-proaches to infer the mechanisms of action of synergistic combinations within cancer signaling pathways. For example, Liu et al. proposed TranSynergy, a drug synergy prediction model that uses the interaction between drug target genes in a protein-protein interaction (PPI) network (Liu and Xie, 2021). However, TranSynergy only relies on target gene information, neglecting information on upstream and downstream activities of the targets and their differential contributions to synergy. More recently, Tang et al. developed SynPathy, a deep learning model for drug synergy prediction that incorporates drug-pathway associations (Tang and Gottlieb, 2022). SynPathy calculates pathway enrichment scores as a measure of the distance between target genes of each drug in a combination and pathway genes in the PPI network. These pathway enrichment scores, along with chemical structure information, are then combined to fit the model and infer pathway importance scores for each combination. More recently, Wu et al. introduced ForSyn (Wu et al., 2023), a deep forest-based method. Although ForSyn implemented a gene enrichment analysis to identify cancer-related pathways, it does not directly identify them through prediction.

Here we present a Drug synergy Interaction Prediction (DIPx) based on tumor- and drug-specific pathway activation scores (PASs). PASs are biologically motivated features that provide potentially relevant information on the underlying mechanisms of synergistic combinations. We trained and validated DIPx using the AstraZeneca-Sanger (AZS) DREAM Challenge dataset (Menden et al., 2019), and compared its performance with the best performing method in the Challenge. Furthermore, we assessed the generalizability of the model by validating it on the ONeil dataset (O’Neil et al., 2016), and provided illustrations of pathways that could mediate the synergistic combinations found by DIPx. DIPx is publicly available at https://www.github.com/tracquangthinh/DIPx.

Results

A pathway based drug synergy prediction model

Figure 1 provides an overview of DIPx, which uses gene expression, mutation profiles, gene-interaction network and drug synergy data to generate PAS of upstream, downstream and driver genes. Based on the PAS combination of these gene sets, we train a random forest prediction model for DIPx and validate its predictive performance in the AZS dataset; see, e.g., Ishwaran and Lu (2019) for a more detailed description of the random forest model. For a new experiment (drug A + drug B, cell line C), DIPx provides the predicted Loewe score and a confidence score. The test set consisted of two subsets: (i) Test Set 1 includes combinations from the training set, and (ii) Test Set 2 includes combinations absent from the training set. Together, both sets assess the generalizability of the prediction for new patients and new combinations. The analysis involved a total of 75 cell lines tested in 910 combinations in the AZS dataset. DIPx was also validated using an external dataset, as shown in Figure 1a.

Overview of DIPx. a) The AZS omics data were used to train and validate the model. The test set was split into two subsets: Test Set 1 contained combinations found in the training set, while Set 2 comprised combinations not found in the training set. The model was also externally validated using the ONeil dataset. b) A cartoon illustration of the ERBB pathway in a breast cancer cell line treated with the combination of Capivasertib + Sapitinib. Capivasertib targets the AKT gene, whereas Sapitinib targets the ERBB genes. Pathway genes were classified into upstream and downstream genes relative to the position of the target genes in the network. c) The drug synergy prediction model was trained using pathway activation scores (PAS) of the upstream, downstream and driver genes. d) The predicted and observed Loewe scores of a cell line achieved the median Spearman correlation in Test Set 1 of the AZS dataset. The color of each bar shows the confidence score information with the threshold of 0.75. e) The main pathways that contribute to the prediction of the synergy of the Capivasertib + Sapitinib combination. f) The proportion of validated high synergistic predictions (Loewe score ≥ 20) increases with higher confidence scores. The x-axis presents four groups defined by quartiles of confidence scores. This figure was created using BioRender.com.

Figure 1b illustrates the ERBB signaling pathway in relation to the Capivasertib + Sapitinib combination, where the genes belonging to the pathway are classified into upstream and downstream genes relative to the position of the target genes: AKT targeted by Capivasertib and ERBB targeted by Sapitinib. Putative driver mutations were identified in each sample based on a wellcharacterized list of frequently mutated genes in cancer; see Section 4.3. We first calculate the PAS of the upstream and downstream part of the pathway relative to the driver genes; see the Methods section for details. The PAS values are then combined to train a random forest regression model. Given a new drug combination experiment, DIPx predicts the Loewe score for drug synergy, as shown in Figure 1c.

Figure 1d presents the predicted and observed synergies for the SW900 lung cancer cell line, which has a median correlation of 0.50 among the cell lines in Test Set 1; each bar in the figure represents a drug combination. The best predicted combinations include BCL2L1 + AZD5582, AZD5582 + etoposide and doxorubicin + AZD5582, with predicted Loewe scores of 42.34, 26.60, and 25.72, respectively, and high confidence scores of 1.0, 0.90, and 0.82, respectively. A combination with Loewe score greater than 20 is considered highly synergistic (Menden et al., 2019). Although the combination of doxorubicin + AZ12623380 is predicted to have high synergy, it is a low confidence prediction with a confidence score of 0.33. Indeed, the observed Loewe synergy score for this combination is near zero.

The use of PAS allows DIPx to infer the potential biological mechanisms of synergistic drug combinations. Figure 1e shows pathways with the highest contribution to prediction of drug synergy of the Capivasertib + Sapitinib combination: these include the ERBB-related pathways (ERBB2 signaling pathway, ERBB pathway), and tumor-related pathway (lymph-node metastates, focal adhesion).

Figure 1f demonstrates the association between the confidence scores and the validation of predictions. The x-axis represents the confidence scores grouped into quartiles, while the y-axis displays the proportion of validated high synergy (Loewe score ≥ 20). Predictions with higher confidence scores are expected to exhibit a greater level of validation. Indeed, in this figure, the proportion of high synergistic predictions that are validated in the combination of Test Set 1 and 2 of the AZS dataset increases as the confidence score rises.

Validation and comparisons in the AZS dataset

We evaluated the performance of DIPx in the AZS test sets and compared it with TAIJI, which was the best performing method in the AZS DREAM Challenge (Li et al., 2018). TAIJI was trained using both monotherapy drug-response and molecular data. Since DIPx uses only molecular data, to make a fair comparison, we trained TAIJI using only molecular features and referred to it as TAIJI-M.

Figure 2a shows the correlation between the predicted and observed Loewe scores of 963 experiments in Test Set 1 (r = 0.5, 95% CI: 0.47–0.53), where each experiment represents a combination drug A + drug B tried on cell line C, yielding one data point. In comparison, TAIJI-M gives r = 0.38 (95% CI: 0.34–0.42). We also bootstrapped the training set (n = 100 times) and for each bootstrap replicate calculated the Spearman correlation between the predicted and observed scores of all experiments. As illustrated in Figure 2b, DIPx achieved stable Spearman correlations across all bootstrap replicates, which are significantly higher than that of TAIJI-M. The bootstrap distribution actually indicates that the Spearman correlation from DIPx is negatively biased, while from TAIJI it is slightly positively biased. This means that the gap between the bias-corrected estimates of the Spearman correlations from DIPx and TAIJI-M would be even larger; see the Method section for a theoretical explanation.

Performance of DIPx in the AZS dataset. This includes both Test Set 1 (panels a, b, c, f) and Test Set 2 (panels d, e, g).

a) Comparison of predicted vs observed synergy scores for all experiments in Test Set 1. b) Comparison of DIPx vs TAIJI-M in terms of the correlation between predicted and observed synergy scores from all experiments in Test Set 1. Each boxplot shows the results based on 100 bootstrap replicates of the training set. c) Comparison of DIPx and TAIJI-M performance across cell lines in Test Set 1. Each point represents the correlation between the predicted and observed synergy for a given cell line. d) Comparison of DIPx vs TAIJI-M in Test Set 2. Each boxplot displays the correlations between the predicted and observed values obtained from 100 bootstrap replicates of the training set. e) Comparison of performance between DIPx and TAIJI-M in Test Set 2 in relation to the number of drugs in common (x-axis) between the combinations in the test set and the training set. and g) DIPx vs TAIJI-M in three groups classified by monotherapy sensivitity of two drugs in a combination in Test Set 1 (f) and Test Set 2 (g).

Figure 2—figure supplement 1. 10-fold cross validation of DIPx in the training set of the ASZ DREAM dataset

To demonstrate that DIPx does not overfit the training set, we performed a 10-fold cross-validation for DIPx. Figure 2-figure supplement 1 shows the Spearman correlation between the predicted and observed Loewe scores across the ten folds. DIPx achieved a median correlation of 0.48, which is comparable to the correlation of 0.50 in Test Set 1. This indicates that there is no evidence of overfitting in the training set.

Furthermore, we compared the performance of DIPx and TAIJI-M across all cell lines in Test Set 1 using a Spearman correlation between the predicted and observed synergy scores, as shown in Figure 2c. A majority of the cell lines (63%) are below the diagonal line, indicating that DIPx outperforms TAIJI-M in predicting synergy scores for these cell lines.

We also compared the performance of DIPx and TAIJI-M in Test Set 2. As expected, the prediction performance of both methods was worse in Test Set 2 than in Test Set 1 since Test Set 2 consists of new combinations absent from the training set. The Spearman correlation of the observed vs predicted synergy using DIPx is 0.26 (95% CI: 0.22–0.30), which is greater than 0.18 (95% CI: 0.16–0.20) using TAIJI-M. However, the difference is not statically significant. Figure 2d show that this result is stable across 100 bootstrap replications of the training set. A similar downward bias for DIPx is observed in the bootstrap distribution.

To investigate the effect of unseen combinations on prediction performance, we divided each combination (drug A + drug B) in Test Set 2 into one of three groups based on the number of individual drugs present in the training set: (i) neither drug A nor drug B in the training set (‘no drug’), (ii) either drug A or drug B in the training set (‘one drug’), (iii) and both drugs A and B in the training set (“two drugs”), as shown in Figure 2e. Overall, both DIPx and TAIJI-M showed improved performance as the number of drugs present in the training set increased. For experiments in which both drugs were not in the training set (n = 262), TAIJI-M achieved a median correlation of 0.11, while DIPx performed worse with a median correlation of –0.03. For experiments with at least one drug in the training set (n = 2, 499), both methods showed improved performance with median correlations of 0.16 and 0.12 for DIPx and TAIJI-M, respectively. When both drugs in an experiment were present in the training set (n = 4, 370), DIPx achieved a median correlation of 0.30, which was better than TAIJI-M”s performance (r = 0.22, p-value < 6 × 10−4).

Monotherapy drug response profiles have been shown to correlate with synergistic effects and contribute to improving prediction performance, e.g., in TAIJI (Li et al., 2018). Here, we compared the performance of DIPx and TAIJI-M in relation to monotherapy sensitivity as measured by the IC50 value. We categorized each experiment in the AZS test sets into three groups according to the monotherapy response. Briefly, we first calculated the median sensitivity to monotherapy for each drug A (TA) across all experiments. Measuring the response of a cell line to drug A in an experiment by SA, the drug is considered sensitive if SA ≥ TA. We then compared the synergy of a combination of drugs A and B in relation to the monotherapy sensitivity to both drugs, only one drug, or neither drug.

In Test Set 1, we observe some improvement by DIPx in all three groups of monotherapy sensitivity, with the highest performance in the group sensitive to both drugs (median r = 0.48), but they are not statistically significant, see Figure 2f. In Test Set 2, TAIJI-M and DIPx performed similar in the group with no sensitive drug (median r = 0.21 vs r = 0.20 by DIPx, p-value ∼ 0.68). Interestingly, we found that, while the performance of DIPx improved as the number of monotherapy-sensitive drugs in a combination increased, the performance of TAIJI-M decreased, see Figure 2g. All prediction results are provided in Supplementary Table S1.

External validation of DIPx in the ONeil dataset

We used a similar computational approach to evaluate the prediction performance of DIPx in relation to the sensitivity of the constituent monotherapies and the impact of unseen combinations in the ONeil dataset. As shown in Figure 3a, the performance of DIPx improved with an increasing number of monotherapy-sensitive drugs in the combination, consistent with the results of Test Set 2 of the AZS data. The highest Spearman correlation between the predicted and observed scores was seen in combinations with two sensitive drugs (median r = 0.11). In relation to the number of drugs in a combination present in the training set, DIPx achieved better performance for com-binations with none or one drug in the training set (middle box plot, Figure 3 - figure supplement 2).

Prediction performance of DIPx in the ONeil dataset.

a) monotherapy sensitivity, b) 29 cell lines from 6 cancer tissues. The y-axis in all box plots shows the Spearman correlation between predicted and observed values in 100 bootstrap replicates.

Figure 3—figure supplement 2. Prediction performance of DIPx in the ONeil dataset, grouped by unseen combinations.

TAIJI-M relies on a gene-gene interaction network to calculate post-treatment gene expression. This approach limits its applicability to new datasets, as TAIJI-M can only predict synergy scores for drug combinations present in the training dataset. Among the set of drug combinations with both drugs present in the training set, both DIPx and TAIJI-M perform poorly, with Spearman correlations between predicted and observed synergy scores of 0.09 and 0.05, respectively. The poor performance could be due to the limited number of drug combinations (42/583).

We also analyzed the prediction performance of DIPx in the 29 cell lines from 6 different cancer tissues (Figure 3b). Colon cancer (yellow boxplots) and lung cancer cell lines (purple boxplots) showed better validation compared to cell lines from breast, ovarian, melanoma, and prostate cancers.

Inference of the mechanism of action based on PAS

The use of PAS in DIPx allows us to infer the potential mechanisms of action of drug combinations while maintaining the prediction performance of the model. For instance, in Test Set 1 of the AZS data, DIPx suggests the involvement of ERBB2 signaling pathways in the Capivasertib + Sapitinib combination, as illustrated by the top pathways depicted in Figure 1e and marked yellow in Figure 4a. This combination therapy has shown promise in overcoming resistance to anti-ERBB2 monotherapy in HER2+ breast cancer (Fujimoto et al., 2020), and ERBB2 has been identified as a key biomarker associated with synergistic responses to this combination in the AZS DREAM Challenge study (Menden et al., 2019).

Inference of pathway importance scores in the AZS dataset.

a) Scatter plot showing feature importance (x-axis) vs PAS (y-axis) for the Capivasertib + Sapitinib combination. Pathways with high PAS and feature importance (top 5%) are of particular interest. b, c) Top pathways contributing to the prediction of the combinations in Test Set 1 (b) and Test Set 2 (c). For each pathway, the bar plots show its feature importance. d, f) Functional interaction between the pathway vs driver genes (x-axis) and the pathway vs target genes (y-axis) of the top pathways suggested by DIPx in the SW900 cell line treated with synergistic combination BCL2L1 + AZD5582 (d) and the non-synergistic combination Doxorubicin + AZ12623380 (f). The z-score from network enrichment analysis (NEA) is a measure of functional interaction between two gene sets. A higher z-score indicates a stronger interaction compared to a random permutation of the network. The upper right quadrant (z-score > 1.96) represents pathways that are potentially interesting. e, g) Cartoon illustration of the potential pathways mediated by the synergistic combination of BCL2L1 + AZD5582 (e) and the non-synergistic combination Doxorubicin + AZ12623380 (g). This figure was created using BioRender.com.

Figure 4—figure supplement 3. A cartoon illustration of the RAS pathway mediated by the Selumetinib + MK-2206 combination

Figure 4—figure supplement 4. Observed vs predicted inhibition in the SW900 cell line treated by BCL2L1 + AZD5582 and Doxorubicin + AZ12623380 combinations

Figure 4—figure supplement 5. Functional interaction between driver genes, target genes, and top pathways suggested by DIPx in the SW900 cell line treated with BCL2L1 + AZD5582.

Figure 4—figure supplement 6. Functional interaction between driver genes, target genes, and top pathways suggested by DIPx in the SW900 cell line treated with Doxorubicin + AZ12623380.

Figure 4—figure supplement 7. Top pathways contributing to the prediction of the MK2206 + Erlotinib combination in the ONeil dataset

Figure 4a further shows the distribution of feature importance versus PAS for all pathways for Capivasertib + Sapitinib combination. The feature importance value (x-axis) is calculated using the permutation method of Ishwaran and Lu (2019). The PAS value (y-axis) represents the median PAS across samples treated with this combination in two test sets. Our focus is on pathways with high feature importance (e.g., the top 5%) as well as highly activated (top 5% PAS). Therefore, the top-right section of Figure 4a is the interesting region. We present additional examples to further demonstrate the capabilities of DIPx. Figure 4b gives the top pathways of MEDI3622, an ADAM17 inhibitor, in combinations with AKT inhibitors including Capivasertib and MK-2206. These ADAM17 + AKT combinations target multiple parts of the PI3K/AKT pathway through ERBB activation (Menden et al., 2019), which aligns with the potential pathway candidates suggested by DIPx.

One of the key strengths of DIPx is its ability to infer potential mechanisms of both known and novel drug combinations, even in cases where limited biological or clinical information is available. This capability is particularly valuable for new combinations that have not been included in the training set. For instance, in Figure 4c, we present the key pathways identified for the Selumetinib + MK-2206 combination from Test Set 2 of the AZS data. We observe the involvement of RAS signaling, with Selumetinib targeting MEK and MK-2206 targeting AKT, as shown in Figure 4-figure supplement 3. A recent clinical study has used Selumetinib + MK-2206 to target downstream components of the RAS pathway (Chung et al., 2017).

If the drugs in a combination have the same target, the efficacy of the combination is likely similar to that of each individual drug at higher doses, i.e., they will only have an additive effect. So it seems reasonable to hypothesize that a synergistic combination is more likely to occur when the two drugs have different targets (Chen et al., 2015). But how should the targets be related to each other? To investigate this, we examine the pathways suggested by DIPx. First, we choose a synergistic combination of BCL2L1 + AZD5582 in the SW900 cell line for further illustration. The contour plot of the BCL2L1 + AZD5582 inhibition in the SW900 cell line is illustrated in Figure 4 - figure supplement 4a. We first collected the top 15 pathways (ranked by feature importance) for this BCL2L1 + AZD5582 combination suggested by DIPx. The full list of these pathways is shown in Figure 4 - figure supplement 5. Figure 4d illustrates the functional interaction between the genes of the top 15 pathways and the driver genes of the SW900 cell line (x-axis); the target genes of the combination BCL2L1 + AZD5582 (y-axis). To assess the strength of this interaction, we used the network enrichment analysis (NEA) (Alexeyenko et al., 2012), which provides z-score, an enrichment score, indicating the degree of interaction. A higher z-score reflects a stronger interaction between the two gene sets. The top pathways exhibiting high functional interaction with both the driver genes and target genes (z-score > 1.96) are particularly notable, located in the upper right quadrant of Figure 4d. In particular, the apoptosis pathway via NF-kB (highlighted in green) has the highest pathway-target interaction among these pathways. Figure 4e shows the cartoon illustration of the pathway in which the drug BCL2L1 targets BCL-xL and AZD5582 targets XIAP. This suggests an explanation for the observed synergy between the two drugs. Thus, it appears that in this case we get synergy when the drugs target different parts of a driving pathway, either directly or via other functional interactions.

As a negative control, we examine the non-synergistic combination Doxorubicin + AZ12623380, which targets the same gene TOP2; see Figure 4f and g and Figure 4 - figure supplement 4b. We similarly obtain 15 top-ranking pathways according to feature importance, but now we do not expect to see anything obviously relevant to the SW900 cell line (more details in Figure 4 - figure supplement 6). Some pathways that have a high functional interaction with the target genes (upper-left quadrant) have little interaction with the drivers. There are no clearly outlying points in the upper-right quadrant; the two pathways near the boundary are (i) Shen_Smarca2_targets_up, containing genes whose expression negatively correlated with the expression of the SMARCA2 gene in prostate cancer samples, discovered in relation to androgen-induced proliferation in the prostate; and (ii) Kokkinakis_Methione _deprivation_48hr_up, which contains up-regulated genes in melanoma cell-line MEWO cells after 48h of methionine deprivation. They do not appear to be relevant for the lung cancer cell line SW900.

We also applied DIPx to identify potential activated pathways in the ONeil dataset. Figure 4 -figure supplementary 7 highlights the key pathways contributing to the MK2206 + Erlotinib com-bination. The most significant pathway is ‘Metabolism by CYP Enzymes.’ Previous studies (Molife et al., 2014) have reported that both MK2206 and Erlotinib are metabolized by the CYP enzyme family, further supporting this finding.

PAS captures the functional interaction of drug targets

In Figure 5a, using the AZS data, we compare the observed drug synergy of combinations of two drugs that share some target genes vs those that do not share any target genes. No significant differences were observed (p-value > 0.72), suggesting that non-overlapping drugs in terms of their targets do not necessarily result in improved drug synergy.

a) Comparison of drug synergy between combinations (drug A + drug B) with vs without overlapping target genes. The numbers in parentheses show the sample sizes of each group. b) Drug synergy between four groups in relation to increased functional interaction between the target genes of the two drugs. c) Comparison between the observed functional interaction (z-score in the network enrichment analysis) and the predicted z-score by PAS.

However, we also observed synergy when the two drugs target different genes in the same pathway. More generally, we hypothesize that synergistic effects occur when the targets have functional interaction. As before, the functional interaction is assessed using NEA (Alexeyenko et al., 2012), where a higher z-score value indicates a stronger functional interaction between the two drugs. Figure 5b shows the observed drug synergy (y-axis) in the AZS data for the four groups defined by the quartile values of the z-scores (x-axis). It indicates that combinations with higher functional interaction are more likely to achieve higher drug synergy, with the highest z-score group (z ∈ (2.97, 29.3]) exhibiting the most favorable drug synergy (median Loewe score = 10.34).

However, when added to the prediction model, the functional interaction z-score did not improve the prediction of synergy (data not shown). Statistically, this can happen if PAS already captures the functional interaction information. To show this, using the AZS training data, we trained a prediction model using PAS as the feature and the functional interaction z-score as the output. We then evaluated the performance of the model in the test set. As shown in Figure 5c, we observed a significant correlation between the predicted and observed z-scores, with a Spearman correlation coefficient of 0.46. This explains why the functional interaction does not give additional predictive power in our model. All medians of predicted and observed Loewe scores related to Figure 5c are provided in Supplementary Table S2.

Discussion

We have developed and validated DIPx, an advanced computational model that incorporates gene expression and mutation profiles to predict synergistic drug combinations. DIPx performs well against the best performing method in the AstraZeneca-Sanger DREAM Challenge. Through the use of tumor- and patient-specific pathway activation scores, DIPx also provides valuable information on the potential underlying pathways associated with an observed synergistic drug interaction. In addition to rigorous validation using the AZS dataset, DIPx is further validated on the independent ONeil dataset. This comprehensive validation ensures the robustness and reliability of DIPx in predicting drug synergy across different cancers.

We have compared the perfomance of DIPx to TAIJI-M, the molecular-based model of TAIJI (Li et al., 2018). The extra information from the use of monotherapy data in TAIJI is rather small, approximately 10% increase in the overall Spearman correlation (Li et al., 2018), and, of course, we could also use such data in DIPx, but it is more convenient and informative to focus the comparisons on prediction based on molecular data alone. For instance, this allows us to compare DIPx with TAIJI-M on the prediction of combinations that contain untrained drug(s), which is not possible with the full TAIJI.

The recent availability of large-scale drug combination assay data has allowed the development of realistic prediction models for drug synergy. These datasets offer a substantial number of samples encompassing hundreds of combinations, allowing for extensive validation studies. However, it is important to note that these datasets were generated using different protocols and drug screening techniques. For instance, the AZS data used a 5-by-5 concentration matrix, while the study by ONeil et al. used a 4-by-4 format. In addition, there is limited overlap in the cell lines used among the datasets. These differences pose challenges to the proper validation of prediction methods (Menden et al., 2019). Exploring new datasets or applying novel techniques in the training process (e.g., transfer learning) will be our future direction to improve the performance of DIPx.

A particular strength of our study is that we use the best-performing method in the Challenge as a benchmark. This is a convenient and robust benchmarking, as there were 160 teams that participated in the Challenge (73 teams submitted in the final round). Altogether, these teams used practically all of the commonly used machine learning tools; see the summary in Menden et al. (2019). Another strength is our use and validation of the confidence score metric, which captures the statistical uncertainty in the predicted synergy by a single number. This is more convenient for clinical interpretation than the standard prediction interval, because there is a target level for which a combination is considered synergistic, so the score measures our confidence in achieving the target.

Despite promising results, our study has several limitations. First, the use of cell lines as training and validation samples from the AZS and ONeil datasets may not fully capture the heterogeneity present in actual tumors. Second, the computation of PAS relies solely on the primary target genes of the drug combinations, potentially disregarding valuable information from non-primary targets. There could also be off-targets that we do not know about. This limitation might lead to the loss of information about the broader effects of drug combinations. Third, cancer is a heterogeneous disease that occurs in many tissues. Even within a single tissue, cancer exhibits distinct (molecular) subtypes with varying biological mechanisms and clinical outcomes. Since DIPx was developed using pan-cancer datasets, it may not be optimal for tissue-specific predictions. Our future plan for DIPx would be building cancer-specific models.

Last but not least, prediction of previously untrained combinations remains a great challenge. The worst case is for combinations of drugs that were not previously trained, with the Spearman correlation only around 0.1. However, from a clinical perspective, it is perhaps more realistic to look for combinations among drugs previously trained in monotherapy or in other combinations. Improving the prediction for the combination of such drugs would be worthwhile.

Methods and Materials

Pathway activation score for drug combinations

Pathway activation scores (PASs) are the key features in DIPx. The PAS of pathway P in cell line C is calculated for each drug combination (drug A + drug B) and pathway P. Genes in pathway P are grouped into three subgroups: (a) Gu, which includes all the target genes of drugs A and B, as well as the upstream genes of pathway P; (b) Gd , which includes the downstream genes of pathway P; and (c) Gdr, which consists of all the driver genes of cell line C in pathway P. In the example of the ERBB pathway targeted by Capivasertib + Sapitinib. (Figure 1b), Gu consists of ERBB, PI3K, and also AKT; Gd contains MTOR, RAS and MAPK, while Gdr includes TP53 and ERBB2.

The score for upstream activity (PASu) is calculated by the sum of mRNA expression for genes in Gu. Similarly, the scores for the downstream activity (PASd) and the set of driver genes (PASdr) are calculated from Gd and Gdr. In practice, the genes of the N = 4,762 curated human pathways are provided from the MsigDB database (version 6.2) (Liberzon et al., 2015). The target genes of the drugs are collected from the AZS dataset and extended from the DrugBank database (Wishart et al., 2018) and the ChEMBL database (Zdrazil et al., 2024). The extraction of the driver genes of the cell lines is described in the Datasets section.

A pathway based model for drug synergy prediction

The training features of DIPx consist of three components: upstream activity (PASu), downstream activity (PASd), and driver genes (PASdr), as shown in Figure 1b. The final training matrix has a size of K experiments by 14,286 PASs, where each row corresponds to a specific experiment (drug A + drug B, cell line C).

To address potential sparsity in the training matrix caused by pathways with no target or driver genes, we explored an alternative model with N = 4,762 additional features. Each feature corresponds to a pathway P and is calculated as S(g) * (w1 + w2), where S(g) represents the sum of mRNA expression for all genes in pathway P, and w1 and w2 denote the functional interactions between gene sets: (pathway genes ↔ target genes) and (pathway genes ↔ driver genes), respectively. The functional interactions were estimated using NEA and converted into normal probability scores for w1 and w2. The feature value is zero only when the pathway lacks both targets and driver genes, as well as any interactions with drug targets and driver genes. Additionally, we incorporated the NEA enrichment score between target genes and driver genes into the final matrix. Despite adding these new features, the alternative model did not exhibit any significant improvements in predictive power (data not shown).

For the predictor, we used the random forest algorithm implemented in the randomforestRSC package (with default parameters) in R version 4.0.4. During the development of DIPx, we experimented with various machine learning methods, such as the support vector machine (SVM) and the elastic net. However, we found that these other methods yielded comparable results and that tuning their parameters did not significantly improve prediction performance while requiring extensive additional computations (data not shown). The random forest algorithm in the random- forestRSC package also offers multiple options to calculate the importance of features. In this study, we used the permutation (or Breiman-Cutler) method (Ishwaran and Lu, 2019) to infer the importance of each PAS.

The confidence score (CS) is used to assess the statistical quality of synergy prediction; see Section 5.6 in Pawitan (2001) for the confidence concept in general. First, as previously defined for example in (Menden et al., 2019), a combination is considered synergistic if the Loewe score is greater than or equal to 20. For each sample s, we have the actual predicted synergy Ps (the output of the regression random forest model). We then generate Nb = 100 bootstrap replicates of the training data and obtain the bootstrap predictions for the sample: . The CS of Ps is defined as follows:

The bootstrap replicates are also used to evaluate the standard errors (se) of the Spearman correlation between the observed and predicted synergy scores in the test sets. The 95% confidence intervals are computed by the usual formula: , where is the observed Spearman correlation. Though less frequently used, the bootstrap can also be used for bias correction (Pawitan, 2001, Section 5.2). Bias occurs if there is a nontrivial gap between the observed estimate and the mean of the bootstrap replications and bias correction is used to adjust the original estimate . Theoretically, where F is the underlying data distribution. So, the bias-corrected estimate should be

In practice, the bias is estimated by where are the bootstrap replicates of . When the estimated bias is negative, as we ob-served for DIPx, the bias-corrected estimate is shifted upward. And vice versa, if the bias is positive, as observed for TAIJI-M, the corrected estimate is shifted downward.

Computing p-values using the bootstrap

To compare the predictive performance of DIPx and TAIJI-M (e.g., as shown in Figure 2), the boot- strap method can be used to compute a confidence interval for differential correlation in the test set. However, there is a close relationship between p-values and confidence intervals; see Pawitan (2001), Chapter 5; particularly p.134. In this case, we compute the p-value as follows:

For each bootstrap replication, (i) compute the Spearman correlation between the predicted and observed scores in the test set for DIPx and TAIJI-M. Denote this by r1 and r2. (ii) Compute the difference in the Spearman correlations d = (r1 − r2).

Repeat the bootstrap n = 100 times.

Compute the minimum of these two proportions: proportion of d<0 or proportion of d>0. To overcome the limited bootstrap sample size, we use the normal approximation in computing the proportions.

The two-sided p-value = 2× the minimum proportion in (3).

Datasets

AstraZeneca-Sanger (AZS) DREAM Challenge dataset

The AZS DREAM Challenge is a rigorous competition in the effort to systematically develop and validate drug synergy prediction methods. Indicating the strong interest in the topic, 160 international teams (Menden et al., 2019) participated in the Challenge. It was organized into two subchallenges: i) Prediction for known (tested) combination and ii) Prediction for unknown (untested) drug combinations. The final dataset comprised 11,576 experiments from 85 cell lines and 910 combinations. The gene expression data of these cell lines was obtained from Affymetrix microarray (Menden et al., 2019). However, to ensure consistency between the AZS dataset and the Oneil dataset (O’Neil et al., 2016) (which did not provide gene expression profiles of cell lines), we utilized gene expression data from the Cancer Cell Line Encyclopedia (CCLE) cohort (Ghandi et al., 2019).

Out of the 85 cell lines, we identified 75 cell lines with available gene expression data in the CCLE cohort, resulting in a total of 10,154 experiments involving 903 combinations used in our study. Supplementary Table S3 shows the list of 75 cell lines. For the validation of the prediction model, the data were split into a training set (n = 2,060) and two test sets (n = 963 and 7,131) according to subchallenges 1 and 2, respectively. The first test set contains experiments from 167 combinations (of 69 single drugs) that are also in the training set. The second test set includes experiments with 736 drug combinations that are not in the training set.

We collected gene expression data of 75 cell lines, measuring the transcript per million (TPM) of 37,222 genes, of the CCLE cohort downloaded from the DepMap Portal (Tsherniak et al., 2017). The gene expression data was logarithmically transformed to the base 2 scale for downstream analysis. Additionally, we obtained potential driver genes for these cell lines, including both mutations and fusion genes, from the DepMap Portal. The portal provides information on mutations in 1,637 protein-coding genes associated with cancer biology in a collection of 1,030 cell lines.

To filter the list of mutations, we focused on those occurring in at least 2.5% of the total cell lines. Subsequently, we extracted the list of mutations specific to the 75 cell lines under investigation. For fusion genes, we focused on those present in the Miltelman database (Mitelman, 2022) and occurring at least twice, considering them as relevant for our analysis. The final list of potential driver genes for the 75 cell lines can be found in Supplementary Table S3. On average, each cell line had a median of 29 potential driver genes.

For the drug synergy data, we used a 5-by-5 concentration matrix provided by the Challenge. Drug synergy values were estimated using the Loewe reference model from Combenefit (Di Veroli et al., 2016).

ONeil dataset

ONeil dataset is a large-scale drug synergy screening dataset from Merck&Co company (O”Neil et al., 2016). A total of 23,062 experiments with 583 unique drug combinations (38 monotherapy drugs) was carried out on 38 cancer cell lines by a 4-by-4 drug concentration matrix. Out of 38 cell lines, we found 29 cell lines with available gene expression data from the DepMap Portal. The detail of 29 cell lines is described in Supplementary Table S4. The gene expression data of 37,222 genes from 29 cell lines, as well as the driver genes of these cell lines, were collected from the DepMap Portal using the same procedure as in the AZS dataset. The original release of this dataset provides only the raw data on drug synergy. Here, we calculated the Loewe synergy score for each experiment using Combenefit (Di Veroli et al., 2016). In total, we obtained 16,907 experiments for 583 combinations in 29 cell lines for further analysis. Drug targets of 38 monotherapy drugs were collected from the DrugBank database (Wishart et al., 2018) and the ChEMBL database (Zdrazil et al., 2024). The original names of all pathways mentioned in the manuscript are listed in Supplementary Table S5.

eLife Assessment

This valuable study presents a deep learning framework for predicting synergistic drug combinations for cancer treatment in the AstraZeneca-Sanger (AZS) DREAM Challenge dataset. However, the evidence on the generalizability of the model is incomplete, as part of the validation seems to be flawed by overfitting, and only a modest correlation between predictions and observations was observed in the second, more independent test set. The reported tool, DIPx, could be of use for personalized drug synergy prediction and exploring the activated pathways related to the effects of drug combinations.

Reviewer #1 (Public review):

The authors introduces DIPx, a deep learning framework for predicting synergistic drug combinations for cancer treatment using the AstraZeneca-Sanger (AZS) DREAM Challenge dataset. While the approach is innovative, I have following concerns and comments, and hopefully will improve the study's rigor and applicability, making it a more powerful tool in real clinical world.

(1) In the abstract: "We trained and validated DIPx in the AstraZeneca-Sanger (AZS) DREAM Challenge dataset using two separate test sets: Test Set 1 comprised the combinations already present in the training set, while Test Set 2 contained combinations absent from the training set, thus indicating the model's ability to handle novel combinations". Test Set 1 comprises combinations already present in the training set, likely leading overfitting issue. The model might show inflated performance metrics on this test set due to prior exposure to these combinations, not accurately reflecting its true predictive power on unknown data, which is crucial for discovering new drug synergies. The testing approach reduces the generalizability of the model's findings to new, untested scenarios.

(2) The model struggles with predicting synergies for drug combinations not included in its training data (showing only Spearman correlation 0.26 in Test Set 2). This limits its potential for discovering new therapeutic strategies. Utilizing techniques such as transfer learning or expanding the training dataset to encompass a wider range of drug pairs could help to address this issue.

(3) The use of pan-cancer datasets, while offering broad applicability, may not be optimal for specific cancer subtypes with distinct biological mechanisms. Developing subtype-specific models or adjusting the current model to account for these differences could improve prediction accuracy for individual cancer types.

(4) Line 127, "Since DIPx uses only molecular data, to make a fair comparison, we trained TAJI using only molecular features and referred to it as TAJI-M.". TAJI was designed to use both monotherapy drug-response and molecular data, and likely won't be able to reach maximum potential if removing monotherapy drug-response from the training model. It would be critical to use the same training datasets and then compare the performances. From the Figure 6 of TAJI's paper (Li et al., 2018, PMID: 30054332) , i.e., the mean Pearson correlation for breast cancer and lung cancer are around 0.5 - 0.6.

The following 2 concerns had been include in the Discussion section which are great:

(1) Training and validating the model using cell lines may not fully capture the heterogeneity and complexity of in vivo tumors. To increase clinical relevance, it would be beneficial to validate the model using primary tumor samples or patient-derived xenografts.

(2) The Pathway Activation Score (PAS) is derived exclusively from primary target genes, potentially overlooking critical interactions involving non-primary targets. Including these secondary effects could enhance the model's predictive accuracy and comprehensiveness.

Comments on revisions:

The authors replied to my concerns but they did not address my comments/concerns. Especially for my concern #1: They trained and validated DIPx in the AstraZeneca-Sanger (AZS) DREAM Challenge dataset using two separate test sets: Test Set 1 comprised the combinations already present in the training set. Therefore, test Set 1 comprises combinations already present in the training set, likely leading overfitting issue but they claimed "There is no danger overfitting here" in their "Author Response" letter.

All my other concerns are unchanged too.

Reviewer #2 (Public review):

Trac, Huang, et al used the AZ Drug Combination Prediction DREAM challenge data to make a new random forest-based model for drug synergy. They make comparisons to the winning method and also show that their model has some predictive capacity for a completely different dataset. They highlight the ability of the model to be interpretable in terms of pathway and target interactions for synergistic effects.

In their revised manuscript, the authors attempt to address the points raised about a comparison to the full TAJI model and showing how molecular can be integrated into DIPx.

(1) Their argument that "Using only molecular data allows for more convenient and intuitive inference of pathway importance compared to integrating multiple data types" is unconvincing. It's not clear how adding a data source here confounds pathway inference. They need to add examples.

(2) They have revised the method of calculating p-values instead of bootstrapping them, so the new numbers appear a lot more meaningful now.

(3) The performance on the O'Neill dataset shows the limitations of their training regime and shows the limits of the model in terms of picking new drug combinations. I would argue that is the very definition of overfitting, not being able to model any combination it has never seen.

Reviewer #3 (Public review):

Summary:

Predicting how two different drugs act together by looking at their specific gene targets and pathways is crucial for understanding the biological significance of drug combinations. This study incorporates drug-specific pathway activation scores (PASs) to estimate synergy scores as one of the key advancements for synergy prediction. The new algorithm, Drug synergy Interaction Prediction (DIPx), developed in this study, uses gene expression, mutation profiles, and drug synergy data to train the model and predict synergy between two drugs. Comprehensive comparisons with another best-performing algorithm, TAIJI-M, highlight the potential of its capabilities.

Strengths:

DIPx uses target and driver genes to elucidate pathway activation scores (PASs) to predict drug synergy. This approach integrates gene expression, mutation profiles, and drug synergy data to capture information about the functional interactions between drug targets, thereby providing a potential biological explanation for the synergistic effects of combined drugs. DIPx's performance was tested using the AstraZeneca-Sanger (AZS) DREAM Challenge dataset, especially in Test Set 1, where the Spearman correlation coefficient between predicted and observed drug synergy was 0.50 (95% CI: 0.47-0.53). DIPx's ability to handle novel combinations, as evidenced by its performance in Test Set 2, indicates its potential for predictions of new and untested drug combinations.

Weaknesses:

While the DIPx algorithm shows promise in predicting drug synergy based on pathway activation scores, it's essential to consider its limitations. One limitation is that the availability of training data for specific drug combinations may influence its predictive capability. Further testing and experimental validation of the predictions in future studies would be necessary to fully assess the algorithm's generalizability and robustness.

Author response:

The following is the authors’ response to the previous reviews

We would like to respond to just one remaining concern from Reviewer 1 and Reviewer 2 regarding a potential overfitting in Test Set 1, which involves combinations already present in the training set. DIPx’s (and TAIJI’s) performance in Test Set 1 is better than in Test Set 2, which involves combinations not present in the training set. Let’s consider two general points to highlight why the improved performance is not the result of overfitting.

(1) Suppose we are testing the e ect of one drug D; the training may involve, for example, selecting an optimal dose. A validated e ect of D in an independent test set is not an overfit, even though we are using the same drug in the training and the test set. Testing one drug is an extreme case, but the same idea holds for any number of drugs. What matters is the independence of the test set.

(2) A prediction model P1 will legitimately perform better than model P2, if P1 uses better or more informative features than P2. The features could be those used directly in the model, but they could also be other observable characteristics not directly used in the model, such as optimal subregions of the feature space. DPIx or TAIJI results indicate that the identity of previously trained combinations is one such informative feature. The set of previously trained combinations corresponds to a subregion of the feature space. DIPx’s prediction performance for known combinations would be expected to follow the results from Test Set 1; we cannot expect that if there is an overfitting issue. Finally, we note that Test Set 1 was established and used in the AstraZeneca Dream Challenge for rigorously testing the prediction of known combinations.

Version 3

Abstract

Abstract

Targeted monotherapies for cancer often fail due to inherent or acquired drug resistance. By aiming at multiple targets simultaneously, drug combinations can produce synergistic interactions that increase drug effectiveness and reduce resistance. Computational models based on the integration of omics data have been used to identify synergistic combinations, but predicting drug synergy remains a challenge. Here, we introduce DIPx, an algorithm for personalized prediction of drug synergy based on biologically motivated tumor- and drug-specific pathway activation scores (PASs). We trained and validated DIPx in the AstraZeneca-Sanger (AZS) DREAM Challenge dataset using two separate test sets: Test Set 1 comprised the combinations already present in the training set, while Test Set 2 contained combinations absent from the training set, thus indicating the model’s ability to handle novel combinations. The Spearman correlation coefficients between predicted and observed drug synergy were 0.50 (95% CI: 0.47–0.53) in Test Set 1 and 0.26 (95% CI: 0.22–0.30) in Test Set 2, compared to 0.38 (95% CI: 0.34–0.42) and 0.18 (95% CI: 0.16–0.20), respectively, for the best performing method in the Challenge. We show evidence that higher synergy is associated with higher functional interaction between the drug targets, and this functional interaction information is captured by PAS. We illustrate the use of PAS to provide a potential biological explanation in terms of activated pathways that mediate the synergistic effects of combined drugs. In summary, DIPx can be a useful tool for personalized prediction of drug synergy and exploration of activated pathways related to the effects of combined drugs.

Introduction

Targeted therapies such as specific inhibitors are the most promising class of cancer drugs, but often fail or achieve only temporary remission due to inherent or acquired resistance. Theoretically, by aiming at multiple targets simultaneously, drug combinations can produce a synergistic interaction that increases drug effectiveness and reduces resistance and the chances of relapse (Medicine, 2017; Pemovska et al., 2018; Plana et al., 2022). This is illustrated in the combination of a BRAF inhibitor dabrafenib with a MEK inhibitor trametinib, which suppresses paradoxical reactivation and resistance observed in patients with BRAF-mutated melanoma treated with dabrafenib alone (Zhong et al., 2022; Banzi et al., 2016). This recently approved combination has been shown to improve progression-free and overall survival rates (Subbiah et al., 2023).

The discovery of effective drug combinations has traditionally relied on expert knowledge and understanding of known biological mechanisms (Li et al., 2015). However, this expert-based approach has limited scope to come up with novel combinations. Furthermore, ideally, novel combinations are clinically tested, but it is practically impossible to test all reasonable combinations in a clinical setting. Computational models of drug synergy have shown some potential for personalized prediction of synergistic combinations (Güvenç Paltun et al., 2021; Wu et al., 2022; Kong et al., 2022). These models are typically based on the integration of patient-specific molecular data, such as mutation profiles, gene expression, and drug response information (Güvenç Paltun et al., 2021). For example, TAIJI, the best performing method in the AstraZeneca-Sanger DREAM Challenge, uses these diverse data types to predict drug synergy (Li et al., 2018). The drug combinations predicted to be effective will expand the therapeutic options while maintaining the same level of adverse effects profile. However, despite the advantages offered by modern machine learning methodologies and the availability of large-scale datasets, prediction of synergistic combinations and validating computational models remains challenging. For example, drug screening protocols often vary across studies, and there is a limited overlap in tested drugs and cell lines, complicating the external validation of these models. Additionally, the reliance on ‘black-box’ machine learning approaches hinders the exploration of underlying molecular mechanisms driving synergistic combinations.

To address this limitation, several studies have introduced statistical and computational approaches to infer the mechanisms of action of synergistic combinations within cancer signaling pathways. For example, Liu et al. proposed TranSynergy, a drug synergy prediction model that uses the interaction between drug target genes in a protein-protein interaction (PPI) network (Liu and Xie, 2021). However, TranSynergy only relies on target gene information, neglecting information on upstream and downstream activities of the targets and their differential contributions to synergy. More recently, Tang et al. developed SynPathy, a deep learning model for drug synergy prediction that incorporates drug-pathway associations (Tang and Gottlieb, 2022). SynPathy calculates pathway enrichment scores as a measure of the distance between target genes of each drug in a combination and pathway genes in the PPI network. These pathway enrichment scores, along with chemical structure information, are then combined to fit the model and infer pathway importance scores for each combination. More recently, Wu et al. introduced ForSyn (Wu et al., 2023), a deep forest-based method. Although ForSyn implemented a gene enrichment analysis to identify cancer-related pathways, it does not directly identify them through prediction.

Here we present a Drug synergy Interaction Prediction (DIPx) based on tumor- and drug-specific pathway activation scores (PASs). PASs are biologically motivated features that provide potentially relevant information on the underlying mechanisms of synergistic combinations. We trained and validated DIPx using the AstraZeneca-Sanger (AZS) DREAM Challenge dataset (Menden et al., 2019), and compared its performance with the best performing method in the Challenge. Furthermore, we assessed the generalizability of the model by validating it on the ONeil dataset (O’Neil et al., 2016), and provided illustrations of pathways that could mediate the synergistic combinations found by DIPx. DIPx is publicly available at https://www.github.com/tracquangthinh/DIPx.

Results

A pathway based drug synergy prediction model

Figure 1 provides an overview of DIPx, which uses gene expression, mutation profiles, gene-interaction network and drug synergy data to generate PAS of upstream, downstream and driver genes. Based on the PAS combination of these gene sets, we train a random forest prediction model for DIPx and validate its predictive performance in the AZS dataset; see, e.g., Ishwaran and Lu (2019) for a more detailed description of the random forest model. For a new experiment (drug A + drug B, cell line C), DIPx provides the predicted Loewe score and a confidence score. The test set consisted of two subsets: (i) Test Set 1 includes combinations from the training set, and (ii) Test Set 2 includes combinations absent from the training set. Together, both sets assess the generalizability of the prediction for new patients and new combinations. The analysis involved a total of 75 cell lines tested in 910 combinations in the AZS dataset. DIPx was also validated using an external dataset, as shown in Figure 1a.

Overview of DIPx.

a) The AZS omics data were used to train and validate the model. The test set was split into two subsets: Test Set 1 contained combinations found in the training set, while Set 2 comprised combinations not found in the training set. The model was also externally validated using the ONeil dataset. b) A cartoon illustration of the ERBB pathway in a breast cancer cell line treated with the combination of Capivasertib + Sapitinib. Capivasertib targets the AKT gene, whereas Sapitinib targets the ERBB genes. Pathway genes were classified into upstream and downstream genes relative to the position of the target genes in the network. c) The drug synergy prediction model was trained using pathway activation scores (PAS) of the upstream, downstream and driver genes. d) The predicted and observed Loewe scores of a cell line achieved the median Spearman correlation in Test Set 1 of the AZS dataset. The color of each bar shows the confidence score information with the threshold of 0.75. e) The main pathways that contribute to the prediction of the synergy of the Capivasertib + Sapitinib combination. f) The proportion of validated high synergistic predictions (Loewe score ≥ 20) increases with higher confidence scores. The x-axis presents four groups defined by quartiles of confidence scores. This figure was created using BioRender.com.

Figure 1b illustrates the ERBB signaling pathway in relation to the Capivasertib + Sapitinib combination, where the genes belonging to the pathway are classified into upstream and downstream genes relative to the position of the target genes: AKT targeted by Capivasertib and ERBB targeted by Sapitinib. Putative driver mutations were identified in each sample based on a wellcharacterized list of frequently mutated genes in cancer; see Section 4.3. We first calculate the PAS of the upstream and downstream part of the pathway relative to the driver genes; see the Methods section for details. The PAS values are then combined to train a random forest regression model. Given a new drug combination experiment, DIPx predicts the Loewe score for drug synergy, as shown in Figure 1c.

Figure 1d presents the predicted and observed synergies for the SW900 lung cancer cell line, which has a median correlation of 0.50 among the cell lines in Test Set 1; each bar in the figure represents a drug combination. The best predicted combinations include BCL2L1 + AZD5582, AZD5582 + etoposide and doxorubicin + AZD5582, with predicted Loewe scores of 42.34, 26.60, and 25.72, respectively, and high confidence scores of 1.0, 0.90, and 0.82, respectively. A combination with Loewe score greater than 20 is considered highly synergistic (Menden et al., 2019). Although the combination of doxorubicin + AZ12623380 is predicted to have high synergy, it is a low confidence prediction with a confidence score of 0.33. Indeed, the observed Loewe synergy score for this combination is near zero.

The use of PAS allows DIPx to infer the potential biological mechanisms of synergistic drug combinations. Figure 1e shows pathways with the highest contribution to prediction of drug synergy of the Capivasertib + Sapitinib combination: these include the ERBB-related pathways (ERBB2 signaling pathway, ERBB pathway), and tumor-related pathway (lymph-node metastates, focal adhesion).

Figure 1f demonstrates the association between the confidence scores and the validation of predictions. The x-axis represents the confidence scores grouped into quartiles, while the y-axis displays the proportion of validated high synergy (Loewe score ≥ 20). Predictions with higher confidence scores are expected to exhibit a greater level of validation. Indeed, in this figure, the proportion of high synergistic predictions that are validated in the combination of Test Set 1 and 2 of the AZS dataset increases as the confidence score rises.

Validation and comparisons in the AZS dataset

We evaluated the performance of DIPx in the AZS test sets and compared it with TAIJI, which was the best performing method in the AZS DREAM Challenge (Li et al., 2018). TAIJI was trained using both monotherapy drug-response and molecular data. Since DIPx uses only molecular data, to make a fair comparison, we trained TAIJI using only molecular features and referred to it as TAIJI-M.

Figure 2a shows the correlation between the predicted and observed Loewe scores of 963 experiments in Test Set 1 (r = 0.5, 95% CI: 0.47–0.53), where each experiment represents a combination drug A + drug B tried on cell line C, yielding one data point. In comparison, TAIJI-M gives r = 0.38 (95% CI: 0.34–0.42). We also bootstrapped the training set (n = 100 times) and for each bootstrap replicate calculated the Spearman correlation between the predicted and observed scores of all experiments. As illustrated in Figure 2b, DIPx achieved stable Spearman correlations across all bootstrap replicates, which are significantly higher than that of TAIJI-M. The bootstrap distribution actually indicates that the Spearman correlation from DIPx is negatively biased, while from TAIJI it is slightly positively biased. This means that the gap between the bias-corrected estimates of the Spearman correlations from DIPx and TAIJI-M would be even larger; see the Method section for a theoretical explanation.

Performance of DIPx in the AZS dataset.

This includes both Test Set 1 (panels a, b, c, f) and Test Set 2 (panels d, e, g). a) Comparison of predicted vs observed synergy scores for all experiments in Test Set 1. b) Comparison of DIPx vs TAIJI-M in terms of the correlation between predicted and observed synergy scores from all experiments in Test Set 1. Each boxplot shows the results based on 100 bootstrap replicates of the training set. c) Comparison of DIPx and TAIJI-M performance across cell lines in Test Set 1. Each point represents the correlation between the predicted and observed synergy for a given cell line. d) Comparison of DIPx vs TAIJI-M in Test Set 2. Each boxplot displays the correlations between the predicted and observed values obtained from 100 bootstrap replicates of the training set. e) Comparison of performance between DIPx and TAIJI-M in Test Set 2 in relation to the number of drugs in common (x-axis) between the combinations in the test set and the training set. f) and g) DIPx vs TAIJI-M in three groups classified by monotherapy sensivitity of two drugs in a combination in Test Set 1 (f) and Test Set 2 (g).

Figure 2—figure supplement 1. 10-fold cross validation of DIPx in the training set of the ASZ DREAM dataset

To demonstrate that DIPx does not overfit the training set, we performed a 10-fold cross-validation for DIPx. Figure 2-figure supplement 1 shows the Spearman correlation between the predicted and observed Loewe scores across the ten folds. DIPx achieved a median correlation of 0.48, which is comparable to the correlation of 0.50 in Test Set 1. This indicates that there is no evidence of overfitting in the training set.

Furthermore, we compared the performance of DIPx and TAIJI-M across all cell lines in Test Set 1 using a Spearman correlation between the predicted and observed synergy scores, as shown in Figure 2c. A majority of the cell lines (63%) are below the diagonal line, indicating that DIPx outperforms TAIJI-M in predicting synergy scores for these cell lines.

We also compared the performance of DIPx and TAIJI-M in Test Set 2. As expected, the prediction performance of both methods was worse in Test Set 2 than in Test Set 1 since Test Set 2 consists of new combinations absent from the training set. The Spearman correlation of the observed vs predicted synergy using DIPx is 0.26 (95% CI: 0.22–0.30), which is greater than 0.18 (95% CI: 0.16–0.20) using TAIJI-M. However, the difference is not statistically significant. Figure 2d show that this result is stable across 100 bootstrap replications of the training set. A similar downward bias for DIPx is observed in the bootstrap distribution.

To investigate the effect of unseen combinations on prediction performance, we divided each combination (drug A + drug B) in Test Set 2 into one of three groups based on the number of individual drugs present in the training set: (i) neither drug A nor drug B in the training set (“no drug”), (ii) either drug A or drug B in the training set (‘one drug’), (iii) and both drugs A and B in the training set (‘two drugs’), as shown in Figure 2e. Overall, both DIPx and TAIJI-M showed improved performance as the number of drugs present in the training set increased. For experiments in which both drugs were not in the training set (n = 262), TAIJI-M achieved a median correlation of 0.11, while DIPx performed worse with a median correlation of –0.03. For experiments with at least one drug in the training set (n = 2, 499), both methods showed improved performance with median correlations of 0.16 and 0.12 for DIPx and TAIJI-M, respectively. When both drugs in an experiment were present in the training set (n = 4, 370), DIPx achieved a median correlation of 0.30, which was better than TAIJI-M’s performance (r = 0.22, p-value < 6 × 10−4).

Monotherapy drug response profiles have been shown to correlate with synergistic effects and contribute to improving prediction performance, e.g., in TAIJI (Li et al., 2018). Here, we compared the performance of DIPx and TAIJI-M in relation to monotherapy sensitivity as measured by the IC50 value. We categorized each experiment in the AZS test sets into three groups according to the monotherapy response. Briefly, we first calculated the median sensitivity to monotherapy for each drug A (TA) across all experiments. Measuring the response of a cell line to drug A in an experiment by SA, the drug is considered sensitive if SA ≥ TA. We then compared the synergy of a combination of drugs A and B in relation to the monotherapy sensitivity to both drugs, only one drug, or neither drug.

In Test Set 1, we observe some improvement by DIPx in all three groups of monotherapy sensitivity, with the highest performance in the group sensitive to both drugs (median r = 0.48), but they are not statistically significant, see Figure 2f. In Test Set 2, TAIJI-M and DIPx performed similar in the group with no sensitive drug (median r = 0.21 vs r = 0.20 by DIPx, p-value ∼ 0.68). Interestingly, we found that, while the performance of DIPx improved as the number of monotherapy-sensitive drugs in a combination increased, the performance of TAIJI-M decreased, see Figure 2g. All prediction results are provided in Supplementary Table S1.

External validation of DIPx in the ONeil dataset

We used a similar computational approach to evaluate the prediction performance of DIPx in relation to the sensitivity of the constituent monotherapies and the impact of unseen combinations in the ONeil dataset. As shown in Figure 3a, the performance of DIPx improved with an increasing number of monotherapy-sensitive drugs in the combination, consistent with the results of Test Set 2 of the AZS data. The highest Spearman correlation between the predicted and observed scores was seen in combinations with two sensitive drugs (median r = 0.11). In relation to the number of drugs in a combination present in the training set, DIPx achieved better performance for combinations with none or one drug in the training set (middle box plot, Figure 3-figure supplement 2).

Prediction performance of DIPx in the ONeil dataset.

a) monotherapy sensitivity, b) 29 cell lines from 6 cancer tissues. The y-axis in all box plots shows the Spearman correlation between predicted and observed values in 100 bootstrap replicates.

Figure 3—figure supplement 2. Prediction performance of DIPx in the ONeil dataset, grouped by unseen combinations.

We also obtained TAIJI-M’s results in the O’Neil dataset. The original version of TAIJI-M uses gene expression, CNV, mutation, and methylation data. However, due to the lack of methylation data in the ONeil dataset, we retrained TAIJI-M by excluding the methylation features. According to the final report of TAIJI in the challenge (https://www.synapse.org/Synapse:syn5614689/wiki/396206), Guan et al. reported that methylation features do not contribute to prediction performance in the post-challenge analysis. This means that retraining TAIJI-M without methylation data will not affect the comparison between DIPx and TAIJI-M on the ONeil dataset.

TAIJI-M relies on a gene-gene interaction network to calculate post-treatment gene expression. This approach limits its applicability to new datasets, as TAIJI-M can only predict synergy scores for drug combinations present in the training dataset. Among the set of drug combinations with both drugs present in the training set, both DIPx and TAIJI-M perform poorly, with Spearman correlations between predicted and observed synergy scores of 0.09 and 0.05, respectively. The poor performance could be due to the limited number of drug combinations (42/583).

We also analyzed the prediction performance of DIPx in the 29 cell lines from 6 different cancer tissues (Figure 3b). Colon cancer (yellow boxplots) and lung cancer cell lines (purple boxplots) showed better validation compared to cell lines from breast, ovarian, melanoma, and prostate cancers.

Inference of the mechanism of action based on PAS

The use of PAS in DIPx allows us to infer the potential mechanisms of action of drug combinations while maintaining the prediction performance of the model. For instance, in Test Set 1 of the AZS data, DIPx suggests the involvement of ERBB2 signaling pathways in the Capivasertib + Sapitinib combination, as illustrated by the top pathways depicted in Figure 1e and marked yellow in Figure 4a. This combination therapy has shown promise in overcoming resistance to anti-ERBB2 monotherapy in HER2+ breast cancer (Fujimoto et al., 2020), and ERBB2 has been identified as a key biomarker associated with synergistic responses to this combination in the AZS DREAM Challenge study (Menden et al., 2019).

Inference of pathway importance scores in the AZS dataset.

a) Scatter plot showing feature importance (x-axis) vs PAS (y-axis) for the Capivasertib + Sapitinib combination. Pathways with high PAS and feature importance (top 5%) are of particular interest. b, c) Top pathways contributing to the prediction of the combinations in Test Set 1 (b) and Test Set 2 (c). For each pathway, the bar plots show its feature importance. d, f) Functional interaction between the pathway vs driver genes (x-axis) and the pathway vs target genes (y-axis) of the top pathways suggested by DIPx in the SW900 cell line treated with synergistic combination BCL2L1 + AZD5582 (d) and the non-synergistic combination Doxorubicin + AZ12623380 (f). The z-score from network enrichment analysis (NEA) is a measure of functional interaction between two gene sets. A higher z-score indicates a stronger interaction compared to a random permutation of the network. The upper right quadrant (z-score > 1.96) represents pathways that are potentially interesting. e, g) Cartoon illustration of the potential pathways mediated by the synergistic combination of BCL2L1 + AZD5582 (e) and the non-synergistic combination Doxorubicin + AZ12623380 (g). This figure was created using BioRender.com.

Figure 4—figure supplement 3. A cartoon illustration of the RAS pathway mediated by the Selumetinib + MK-2206 combination

Figure 4—figure supplement 4. Observed vs predicted inhibition in the SW900 cell line treated by BCL2L1 + AZD5582 and Doxorubicin + AZ12623380 combinations

Figure 4—figure supplement 5. Functional interaction between driver genes, target genes, and top pathways suggested by DIPx in the SW900 cell line treated with BCL2L1 + AZD5582.

Figure 4—figure supplement 6. Functional interaction between driver genes, target genes, and top pathways suggested by DIPx in the SW900 cell line treated with Doxorubicin + AZ12623380.

Figure 4—figure supplement 7. Top pathways contributing to the prediction of the MK2206 + Erlotinib combination in the ONeil dataset

a) Comparison of drug synergy between combinations (drug A + drug B) with vs without overlapping target genes. The numbers in parentheses show the sample sizes of each group. b) Drug synergy between four groups in relation to increased functional interaction between the target genes of the two drugs. c) Comparison between the observed functional interaction (z-score in the network enrichment analysis) and the predicted z-score by PAS.

Figure 4a further shows the distribution of feature importance versus PAS for all pathways for Capivasertib + Sapitinib combination. The feature importance value (x-axis) is calculated using the permutation method of Ishwaran and Lu (2019). The PAS value (y-axis) represents the median PAS across samples treated with this combination in two test sets. Our focus is on pathways with high feature importance (e.g., the top 5%) as well as highly activated (top 5% PAS). Therefore, the top-right section of Figure 4a is the interesting region. We present additional examples to further demonstrate the capabilities of DIPx. Figure 4b gives the top pathways of MEDI3622, an ADAM17 inhibitor, in combinations with AKT inhibitors including Capivasertib and MK-2206. These ADAM17 + AKT combinations target multiple parts of the PI3K/AKT pathway through ERBB activation (Menden et al., 2019), which aligns with the potential pathway candidates suggested by DIPx.

One of the key strengths of DIPx is its ability to infer potential mechanisms of both known and novel drug combinations, even in cases where limited biological or clinical information is available. This capability is particularly valuable for new combinations that have not been included in the training set. For instance, in Figure 4c, we present the key pathways identified for the Selumetinib + MK-2206 combination from Test Set 2 of the AZS data. We observe the involvement of RAS signaling, with Selumetinib targeting MEK and MK-2206 targeting AKT, as shown in Figure 4-figure supplement 3. A recent clinical study has used Selumetinib + MK-2206 to target downstream components of the RAS pathway (Chung et al., 2017).

If the drugs in a combination have the same target, the efficacy of the combination is likely similar to that of each individual drug at higher doses, i.e., they will only have an additive effect. So it seems reasonable to hypothesize that a synergistic combination is more likely to occur when the two drugs have different targets (Chen et al., 2015). But how should the targets be related to each other? To investigate this, we examine the pathways suggested by DIPx. First, we choose a synergistic combination of BCL2L1 + AZD5582 in the SW900 cell line for further illustration. The contour plot of the BCL2L1 + AZD5582 inhibition in the SW900 cell line is illustrated in Figure 4-figure supplement 4a. We first collected the top 15 pathways (ranked by feature importance) for this BCL2L1 + AZD5582 combination suggested by DIPx. The full list of these pathways is shown in Figure 4-figure supplement 5. Figure 4d illustrates the functional interaction between the genes of the top 15 pathways and the driver genes of the SW900 cell line (x-axis); the target genes of the combination BCL2L1 + AZD5582 (y-axis). To assess the strength of this interaction, we used the network enrichment analysis (NEA) (Alexeyenko et al., 2012), which provides z-score, an enrichment score, indicating the degree of interaction. A higher z-score reflects a stronger interaction between the two gene sets. The top pathways exhibiting high functional interaction with both the driver genes and target genes (z-score > 1.96) are particularly notable, located in the upper right quadrant of Figure 4d. In particular, the apoptosis pathway via NF-kB (highlighted in green) has the highest pathway-target interaction among these pathways. Figure 4e shows the cartoon illustration of the pathway in which the drug BCL2L1 targets BCL-xL and AZD5582 targets XIAP. This suggests an explanation for the observed synergy between the two drugs. Thus, it appears that in this case we get synergy when the drugs target different parts of a driving pathway, either directly or via other functional interactions.

As a negative control, we examine the non-synergistic combination Doxorubicin + AZ12623380, which targets the same gene TOP2; see Figure 4f and g and Figure 4-figure supplement 4b. We similarly obtain 15 top-ranking pathways according to feature importance, but now we do not expect to see anything obviously relevant to the SW900 cell line (more details in Figure 4-figure supplement 6). Some pathways that have a high functional interaction with the target genes (upper-left quadrant) have little interaction with the drivers. There are no clearly outlying points in the upper-right quadrant; the two pathways near the boundary are (i) Shen_Smarca2_targets_up, containing genes whose expression negatively correlated with the expression of the SMARCA2 gene in prostate cancer samples, discovered in relation to androgen-induced proliferation in the prostate; and (ii) Kokkinakis_Methione _deprivation_48hr_up, which contains up-regulated genes in melanoma cell-line MEWO cells after 48h of methionine deprivation. They do not appear to be relevant for the lung cancer cell line SW900.

We also applied DIPx to identify potential activated pathways in the ONeil dataset. Figure 4-figure supplement 7 highlights the key pathways contributing to the MK2206 + Erlotinib combination. The most significant pathway is ‘Metabolism by CYP Enzymes.’ Previous studies (Molife et al., 2014) have reported that both MK2206 and Erlotinib are metabolized by the CYP enzyme family, further supporting this finding.

PAS captures the functional interaction of drug targets

In Figure 5a, using the AZS data, we compare the observed drug synergy of combinations of two drugs that share some target genes vs those that do not share any target genes. No significant differences were observed (p-value > 0.72), suggesting that non-overlapping drugs in terms of their targets do not necessarily result in improved drug synergy.

However, we also observed synergy when the two drugs target different genes in the same pathway. More generally, we hypothesize that synergistic effects occur when the targets have functional interaction. As before, the functional interaction is assessed using NEA (Alexeyenko et al., 2012), where a higher z-score value indicates a stronger functional interaction between the two drugs. Figure 5b shows the observed drug synergy (y-axis) in the AZS data for the four groups defined by the quartile values of the z-scores (x-axis). It indicates that combinations with higher functional interaction are more likely to achieve higher drug synergy, with the highest z-score group (z ∈ (2.97, 29.3]) exhibiting the most favorable drug synergy (median Loewe score = 10.34).

However, when added to the prediction model, the functional interaction z-score did not improve the prediction of synergy (data not shown). Statistically, this can happen if PAS already captures the functional interaction information. To show this, using the AZS training data, we trained a prediction model using PAS as the feature and the functional interaction z-score as the output. We then evaluated the performance of the model in the test set. As shown in Figure 5c, we observed a significant correlation between the predicted and observed z-scores, with a Spearman correlation coefficient of 0.46. This explains why the functional interaction does not give additional predictive power in our model. All medians of predicted and observed Loewe scores related to Figure 5c are provided in Supplementary Table S2.

Discussion

We have developed and validated DIPx, an advanced computational model that incorporates gene expression and mutation profiles to predict synergistic drug combinations. DIPx performs well against the best performing method in the AstraZeneca-Sanger DREAM Challenge. Through the use of tumor- and patient-specific pathway activation scores, DIPx also provides valuable information on the potential underlying pathways associated with an observed synergistic drug interaction. In addition to rigorous validation using the AZS dataset, DIPx is further validated on the independent ONeil dataset. This comprehensive validation ensures the robustness and reliability of DIPx in predicting drug synergy across different cancers.

We have compared the perfomance of DIPx to TAIJI-M, the molecular-based model of TAIJI (Li et al., 2018). The extra information from the use of monotherapy data in TAIJI is rather small, approximately 10% increase in the overall Spearman correlation (Li et al., 2018), and, of course, we could also use such data in DIPx, but it is more convenient and informative to focus the comparisons on prediction based on molecular data alone. For instance, this allows us to compare DIPx with TAIJI-M on the prediction of combinations that contain untrained drug(s), which is not possible with the full TAIJI.

The recent availability of large-scale drug combination assay data has allowed the development of realistic prediction models for drug synergy. These datasets offer a substantial number of samples encompassing hundreds of combinations, allowing for extensive validation studies. However, it is important to note that these datasets were generated using different protocols and drug screening techniques. For instance, the AZS data used a 5-by-5 concentration matrix, while the study by ONeil et al. used a 4-by-4 format. In addition, there is limited overlap in the cell lines used among the datasets. These differences pose challenges to the proper validation of prediction methods (Menden et al., 2019). Exploring new datasets or applying novel techniques in the training process (e.g., transfer learning) will be our future direction to improve the performance of DIPx.

A particular strength of our study is that we use the best-performing method in the Challenge as a benchmark. This is a convenient and robust benchmarking, as there were 160 teams that participated in the Challenge (73 teams submitted in the final round). Altogether, these teams used practically all of the commonly used machine learning tools; see the summary in Menden et al., 2019. Another strength is our use and validation of the confidence score metric, which captures the statistical uncertainty in the predicted synergy by a single number. This is more convenient for clinical interpretation than the standard prediction interval, because there is a target level for which a combination is considered synergistic, so the score measures our confidence in achieving the target.

Despite promising results, our study has several limitations. First, the use of cell lines as training and validation samples from the AZS and ONeil datasets may not fully capture the heterogeneity present in actual tumors. Second, the computation of PAS relies solely on the primary target genes of the drug combinations, potentially disregarding valuable information from non-primary targets. There could also be off-targets that we do not know about. This limitation might lead to the loss of information about the broader effects of drug combinations. Third, cancer is a heterogeneous disease that occurs in many tissues. Even within a single tissue, cancer exhibits distinct (molecular) subtypes with varying biological mechanisms and clinical outcomes. Since DIPx was developed using pan-cancer datasets, it may not be optimal for tissue-specific predictions. Our future plan for DIPx would be building cancer-specific models.

Last but not least, prediction of previously untrained combinations remains a great challenge. The worst case is for combinations of drugs that were not previously trained, with the Spearman correlation only around 0.1. However, from a clinical perspective, it is perhaps more realistic to look for combinations among drugs previously trained in monotherapy or in other combinations. Improving the prediction for the combination of such drugs would be worthwhile.

Methods and Materials

Pathway activation score for drug combinations

Pathway activation scores (PASs) are the key features in DIPx. The PAS of pathway P in cell line C is calculated for each drug combination (drug A + drug B) and pathway P. Genes in pathway P are grouped into three subgroups: (a) Gu, which includes all the target genes of drugs A and B, as well as the upstream genes of pathway P; (b) Gd, which includes the downstream genes of pathway P; and (c) Gdr, which consists of all the driver genes of cell line C in pathway P. In the example of the ERBB pathway targeted by Capivasertib + Sapitinib. (Figure 1b), Gu consists of ERBB, PI3K, and also AKT; Gd contains MTOR, RAS and MAPK, while Gdr includes TP53 and ERBB2.

The score for upstream activity (PASu) is calculated by the sum of mRNA expression for genes in Gu. Similarly, the scores for the downstream activity (PASd) and the set of driver genes (PASdr) are calculated from Gd and Gdr. In practice, the genes of the N = 4,762 curated human pathways are provided from the MsigDB database (version 6.2) (Liberzon et al., 2015). The target genes of the drugs are collected from the AZS dataset and extended from the DrugBank database (Wishart et al., 2018) and the ChEMBL database (Zdrazil et al., 2024). The extraction of the driver genes of the cell lines is described in the Datasets section.

A pathway based model for drug synergy prediction

The training features of DIPx consist of three components: upstream activity (PASu), downstream activity (PASd), and driver genes (PASdr), as shown in Figure 1b. The final training matrix has a size of K experiments by 14,286 PASs, where each row corresponds to a specific experiment (drug A + drug B, cell line C).

To address potential sparsity in the training matrix caused by pathways with no target or driver genes, we explored an alternative model with N = 4,762 additional features. Each feature corresponds to a pathway P and is calculated as S(g) * (w1 + w2), where S(g) represents the sum of mRNA expression for all genes in pathway P, and w1 and w2 denote the functional interactions between gene sets: (pathway genes ↔ target genes) and (pathway genes ↔ driver genes), respectively. The functional interactions were estimated using NEA and converted into normal probability scores for w1 and w2. The feature value is zero only when the pathway lacks both targets and driver genes, as well as any interactions with drug targets and driver genes. Additionally, we incorporated the NEA enrichment score between target genes and driver genes into the final matrix. Despite adding these new features, the alternative model did not exhibit any significant improvements in predictive power (data not shown).

For the predictor, we used the random forest algorithm implemented in the randomforestRSC package (with default parameters) in R version 4.0.4. During the development of DIPx, we experimented with various machine learning methods, such as the support vector machine (SVM) and the elastic net. However, we found that these other methods yielded comparable results and that tuning their parameters did not significantly improve prediction performance while requiring extensive additional computations (data not shown). The random forest algorithm in the randomforestRSC package also offers multiple options to calculate the importance of features. In this study, we used the permutation (or Breiman-Cutler) method (Ishwaran and Lu, 2019) to infer the importance of each PAS.

The confidence score (CS) is used to assess the statistical quality of synergy prediction; see Section 5.6 in Pawitan (2001) for the confidence concept in general. First, as previously defined for example in (Menden et al., 2019), a combination is considered synergistic if the Loewe score is greater than or equal to 20. For each sample s, we have the actual predicted synergy Ps (the output of the regression random forest model). We then generate Nb = 100 bootstrap replicates of the training data and obtain the bootstrap predictions for the sample: . The CS of Ps is defined as follows:

The bootstrap replicates are also used to evaluate the standard errors (se) of the Spearman correlation between the observed and predicted synergy scores in the test sets. The 95% confidence intervals are computed by the usual formula: , where is the observed Spearman correlation. Though less frequently used, the bootstrap can also be used for bias correction (Pawitan, 2001, Section 5.2). Bias occurs if there is a nontrivial gap between the observed estimate and the mean of the bootstrap replications and bias correction is used to adjust the original estimate. Theoretically, where F is the underlying data distribution. So, the bias-corrected estimate should be

In practice, the bias is estimated by where are the bootstrap replicates of . When the estimated bias is negative, as we observed for DIPx, the bias-corrected estimate is shifted upward. And vice versa, if the bias is positive, as observed for TAIJI-M, the corrected estimate is shifted downward.

Computing p-values using the bootstrap

To compare the predictive performance of DIPx and TAIJI-M (e.g., as shown in Figure 2), the bootstrap method can be used to compute a confidence interval for differential correlation in the test set. However, there is a close relationship between p-values and confidence intervals; see Pawitan (2001), Chapter 5; particularly p.134. In this case, we compute the p-value as follows:

For each bootstrap replication, (i) compute the Spearman correlation between the predicted and observed scores in the test set for DIPx and TAIJI-M. Denote this by r1 and r2. (ii) Compute the difference in the Spearman correlations d = (r1 − r2).

Repeat the bootstrap n = 100 times.

Compute the minimum of these two proportions: proportion of d<0 or proportion of d>0. To overcome the limited bootstrap sample size, we use the normal approximation in computing the proportions.

The two-sided p-value = 2× the minimum proportion in (3).

Datasets

AstraZeneca-Sanger (AZS) DREAM Challenge dataset

The AZS DREAM Challenge is a rigorous competition in the effort to systematically develop and validate drug synergy prediction methods. Indicating the strong interest in the topic, 160 international teams (Menden et al., 2019) participated in the Challenge. It was organized into two subchallenges: i) Prediction for known (tested) combination and ii) Prediction for unknown (untested) drug combinations. The final dataset comprised 11,576 experiments from 85 cell lines and 910 combinations. The gene expression data of these cell lines was obtained from Affymetrix microarray (Menden et al., 2019). However, to ensure consistency between the AZS dataset and the Oneil dataset (O’Neil et al., 2016) (which did not provide gene expression profiles of cell lines), we utilized gene expression data from the Cancer Cell Line Encyclopedia (CCLE) cohort (Ghandi et al., 2019).

Out of the 85 cell lines, we identified 75 cell lines with available gene expression data in the CCLE cohort, resulting in a total of 10,154 experiments involving 903 combinations used in our study. Supplementary Table S3 shows the list of 75 cell lines. For the validation of the prediction model, the data were split into a training set (n = 2,060) and two test sets (n = 963 and 7,131) according to subchallenges 1 and 2, respectively. The first test set contains experiments from 167 combinations (of 69 single drugs) that are also in the training set. The second test set includes experiments with 736 drug combinations that are not in the training set.

We collected gene expression data of 75 cell lines, measuring the transcript per million (TPM) of 37,222 genes, of the CCLE cohort downloaded from the DepMap Portal (Tsherniak et al., 2017). The gene expression data was logarithmically transformed to the base 2 scale for downstream analysis. Additionally, we obtained potential driver genes for these cell lines, including both mutations and fusion genes, from the DepMap Portal. The portal provides information on mutations in 1,637 protein-coding genes associated with cancer biology in a collection of 1,030 cell lines.

To filter the list of mutations, we focused on those occurring in at least 2.5% of the total cell lines. Subsequently, we extracted the list of mutations specific to the 75 cell lines under investigation. For fusion genes, we focused on those present in the Miltelman database (Mitelman, 2022) and occurring at least twice, considering them as relevant for our analysis. The final list of potential driver genes for the 75 cell lines can be found in Supplementary Table S3. On average, each cell line had a median of 29 potential driver genes.

For the drug synergy data, we used a 5-by-5 concentration matrix provided by the Challenge. Drug synergy values were estimated using the Loewe reference model from Combenefit (Di Veroli et al., 2016).

ONeil dataset

ONeil dataset is a large-scale drug synergy screening dataset from Merck&Co company (O’Neil et al., 2016). A total of 23,062 experiments with 583 unique drug combinations (38 monotherapy drugs) was carried out on 38 cancer cell lines by a 4-by-4 drug concentration matrix. Out of 38 cell lines, we found 29 cell lines with available gene expression data from the DepMap Portal. The detail of 29 cell lines is described in Supplementary Table S4. The gene expression data of 37,222 genes from 29 cell lines, as well as the driver genes of these cell lines, were collected from the DepMap Portal using the same procedure as in the AZS dataset. The original release of this dataset provides only the raw data on drug synergy. Here, we calculated the Loewe synergy score for each experiment using Combenefit (Di Veroli et al., 2016). In total, we obtained 16,907 experiments for 583 combinations in 29 cell lines for further analysis. Drug targets of 38 monotherapy drugs were collected from the DrugBank database (Wishart et al., 2018) and the ChEMBL database (Zdrazil et al., 2024). The original names of all pathways mentioned in the manuscript are listed in Supplementary Table S5.